--- 
title: "TMA4315 Generalized Linear Models" 
subtitle: "Compulsory exercise 1: Linear models for Gaussian data" 
date: "Deadline: Friday, September 28, 2018 at 16.00" 
output: # 3rd letter intentation hierarchy
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
  # pdf_document:
  #  toc: false
  #  toc_depth: 2
  

---

```{r setup, include = FALSE}

library(formatR)
showsol <- FALSE
library(knitr)
opts_chunk$set(tidy.opts = list(width.cutoff = 68), tidy = TRUE, warning = FALSE, error = FALSE, message = FALSE, echo = TRUE)

library(INLA)

```


### Problem 3: Compulsory exercise 1

Introduction to the first compulsory exercise by TA Ingeborg - and an introduction to packages and classes in `R`.

The exercise: <https://www.math.ntnu.no/emner/TMA4315/2018h/project1_h18.html>

Packages: `ggplot2`, `gamlss.data`, and so on. Some are already loaded when `R` starts (like `stats`), others must be loaded (like `MASS`).

You are going to make your own package, called `mylm`, which performs multiple linear regression and is a smaller version of `lm`.

**Show how to create package in R Studio**

Classes in R: Something we do not have to think much about, but we use all the time. We are now going to make a new class in R, that we call "test".

```{r}

# takes a word, and returns the index in the alphabet of each letter in an object with class "test"
test <- function(word){
  
  x <- 1:nchar(word)
  y <- match(c(strsplit(tolower(word), "")[[1]]), letters[1:26])
  
  res <- list(x = x, y = y, word = word) # if you are not familiar with lists, you should read up on this
  class(res) <- "test"
  
  return(res)
  
}

```

Now we make an object of this class and try to look at it.

```{r}

myname <- test("Ingeborg")
# "print(myname)" and "myname" returns the same thing in a script, so to simplify we just write "myname" here
myname # prints everything

# lets make a print function that only prints the word
print.test <- function(obj) cat(obj$word)

myname # and now we get only the name

```

Now we want a function that plots objects of this class in a particular way.

```{r}

# important that it is called plot.test with .test at the end!!!
plot.test <- function(obj) plot(obj$x, obj$y, xlab = "Letter", ylab = "Index",
                                main = obj$word, col = rainbow(length(obj$x)), pch = 19, cex = 2)

plot(myname) # we do not have to specify that this is plot.test, because "myname" is already of class "test"!

```

```{r}

# And a summary function
summary.test <- function(obj){
  
  cat("Word: ", obj$word, "\n")
  cat("Length of word: ", length(obj$x), " letters\n")
  cat("Occurrence of each letter:")
  print(table(strsplit(tolower(obj$word), "")))
  
}

summary(myname)

```

Now we have made a class with a plot, print and summary function, and this is what you do in the exercise! But a bit more advanced...

---

Let us look at what happens when we use the plotting function on objects with different classes: The function called `plot`. First we make two new objects that can be plotted:

```{r}

data <- data.frame(x = rnorm(10), y = rnorm(10))
mod <- lm(y ~ x, data = data)

```

And then we plot them:

```{r, echo = c(-1, -3, -5), fig.height = 3}
layout(1)
plot(data)
layout(matrix(1:4, nrow = 1))
plot(mod)
layout(1)
plot(myname)

```

What is happening? `R` reads the class of the objects and uses the plot-function made for that specific class. The user does not have to specify the class as this is already stored in the object!

The different objects we have declared earlier have the following classes:

```{r, results = "hold"}

class(data)
class(mod)
class(myname)

```

---


You will make a new class in `R` called `mylm`, and `R` will also then understand which plot-function to use based on the class.

Note that an object can have more than one class.

---

You write the report using R Markdown, use this template: <https://www.math.ntnu.no/emner/TMA4315/2018h/template_glm.Rmd>

---


### Exercises:

Discuss with the group to get a feeling on what to do in the exercise.

1. Go through how to make an R package together in the group, and make the mylm-package.
2. The core is the `mylm` function. Which formulas are used to
    + calculate parameters estimates?
    + calculate covariance matrix of the estimated regression coefficients?
    + perform type 3 hypothesis tests (remember you need to do the asymptotic normal - so no t-distributions)?
3. You will make `print.mylm`, `plot.mylm` and `summary.mylm`. What should these functions contain?
4. Look at the mylm-template (<https://www.math.ntnu.no/emner/TMA4315/2018h/mylm.R>) and see if you understand it, or if you have questions about some of the parts. In particular, explore the functions `model.frame`, `model.matrix` and `model.response`.
















