---
title: "TMA4315 Generalized linear models H2018"
subtitle: "R test-question for the exam"
author: "Mette Langaas, Department of Mathematical Sciences, NTNU"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: #3rd letter intentation hierarchy
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
#  pdf_document:
  #   toc: true
  #   toc_depth: 2
  #   keep_tex: yes
#  beamer_presentation:
#    keep_tex: yes
#    fig_caption: false
#    latex_engine: xelatex
---

One of the built in datasets in R is called `InsectSprays` and is accessed by writing `InsectSprays`.The data set consists of 72 observations of two variables from different agricultural fields. The variables are:

* `count` of insects remaining after the field was treated with some treatment,
* `spray` the treatment (A-F) that the field was subject to

Fit a generalized linear model with canonical link to the data. Use `count` as the response and the factor `spray` as a covariate. Do not include an intercept term in the model. You may also need to perform additional commands (in addition to fitting the model) on your fitted GLM-object.

Based on your data analysis answer the following questions:

a) Which type of GLM did you fit? Binary, Poisson, Gamma, Normal, other (specify)

b) Which link function did you use? Identity, log, exp, logit, inverse, other (specify)

c) What is the estimated regression coefficient for effect of `spray` of category A: 2.67415

d) What is the estimated variance of the regression coefficient for the effect of `spray` of category C: 0.04

e) Perform a deviance test and report the p-value. 0.006054201

f) What is the conclusion from the deviance test: "The model is good" or "the model is not good"? Not good.

g) Give the estimated covariance between the estimated effect of `spray` B and `spray` F: 0.00 (orthogonal design)

h) Perform an test with $H_0$ stating that the regression coefficient of for spray B is equal to the regression coefficient for spray F, vs. the two-sided alternative. Report the p-value from this test. 

```{r}
fit2=glm(count~spray-1,family="poisson",data=InsectSprays)
summary(fit2)
1-pchisq(fit2$deviance,fit2$df.residual)
vcov(fit2)
C=matrix(c(0,1,0,0,0,-1),ncol=6)
Cbeta=C%*%fit2$coefficients
covCbeta=C%*%vcov(fit2)%*%t(C)
covCbeta
testobs=t(Cbeta)%*%solve(covCbeta)%*%(Cbeta)
testobs
1-pchisq(testobs,1)
# alternatively since orth sprays
testobs2=(fit2$coefficients[2]-fit2$coefficients[6])/sqrt(summary(fit2)$coefficients[2,2]^2+summary(fit2)$coefficients[6,2]^2)
testobs2^2
1-pchisq(testobs2^2,1)
# could also use LRT, but a bit mess to recode
```

```{r,echo=FALSE,eval=FALSE}
# looked at original article and there was also a block factor presen
block=factor(rep(rep(1:6,each=2),6))
cbind(block,InsectSprays$spray,InsectSprays$count)
fit1=glm(count~spray*block,family="poisson",data=InsectSprays)
anova(fit1,fit2)
```


