---
title: "TMA4315 Generalized linear models H2018"
subtitle: "Module 4: Count and continuous positive response data (Poisson and gamma regression)"
author: "Mette Langaas, Department of Mathematical Sciences, NTNU -- with contributions from Ingeborg Hem"
date: " 27.09.2018 and 04.10.2018 [PL], 28.09.2018 and 05.10.2018 [IL]"
output: #3rd letter intentation hierarchy
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    css: "../TMA4315RMarkdown.css"

 # pdf_document:
 #   toc: true
 #   toc_depth: 2
 #   keep_tex: yes
#  beamer_presentation:
#    keep_tex: yes
#    fig_caption: false
#    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,tidy=TRUE,message=FALSE,warning=FALSE,results="hold")
showsol<-TRUE
```

**Hints and reminders are in bold**
 
<span class="question">Questions appear in blue.</span>

<details><summary>Some hints and answers are hidden away in a fold like this</summary>
Well done, you found me!
</details>

The exam style questions are optional, but will be helpful for, well, the exam. They will also help you now to see if you understand the content.

# <a id="interactivew1">Interactive session - first week</a>

## Problem 1: Exam 2005 (Problem 1d-f - slightly modified) - Female crabs and satellites

(Only a subset of 20 crabs in the original data was used in the exam
problems, but we will use the full data set of 173 crabs - so results
will not be the same. Permitted aids for the exam was "all printed and
handwritten material and all calculators" - NB that is not the case for 2018!)

We assume that the number of satellites for each female crab follows a
Poisson distribution and want to perform a Poisson regression using a
log-link to find if there is a connection between the expected number of
satellites `Sa` and the width `W` and colour `C` of the carapace of the
female crab.

* `C`: the color of the female crab (1=light medium, 2=medium, 3=dark medium, 4=dark)
* `W`: width of carapace (cm)

The following model was fitted.

```{r,tidy=FALSE}
crab <- read.table("https://www.math.ntnu.no/emner/TMA4315/2018h/crab.txt")
colnames(crab) <- c("Obs", "C", "S", "W", "Wt", "Sa")
crab <- crab[,-1] #remove column with Obs
crab$C <- as.factor(crab$C)
modelEXAM <- glm(Sa ~ W + C, family = poisson(link = log), data = crab, contrasts = list(C = "contr.sum"))
summary(modelEXAM)
library(car)
Anova(modelEXAM, type = "III", test.statistic = "Wald")
```

<span class="question">a) Write down the estimated model and evaluate the significance of
the estimated coefficients.</span> 

<span class="question">How would you proceed to evaluate the significance of the coefficient for the colour C4. Explain.</span>

<span class="question">Perform a test to evaluate the fit of the model. What is your conclusion?<span class="question">
Use a 5% significance level for your evaluations. 

(In Problem 3d we derive the mathematical formula for the test statistic this test is based on.)



<span class="question">b) Make a sketch *by hand* illustrating the connection between the expected
number of satellites and the width of the carapace for each of the four
colours of the crab. </span>

The estimated multiplicative change in the
expected number of satellites when the width of the carapace is changed
by 1 cm is, according to the model, independent of the colour. 

<span class="question">Explain why.</span>

<span class="question">Also find a 95% confidence interval for this change.</span>


c) Let $\hat{\eta}(x_{W})$ be the estimated linear predictor when the width of the carapace is $x_{W}$ and the crab is "light medium". 

<span class="question">What is the distribution of $\hat{\eta}(x_{W})$? </span>

<span class="question">Which value of $x_W$ would give estimated mean number of satellites equal to 5?</span>

<span class="question">Optional: Using R -- for this value of $x_W$, construct a 95% confidence interval for the mean number of satellites.</span>

<span class="question">Optional: Use `ggplot` to create (and improve) the sketch from b).</span>

---

## Problem 2: Exam 2017 (Problem 1) - Poisson regression

Consider a random variable $Y$. In our course we have considered the univariate exponential family having distribution (probability density function for continuous variables and probability mass function for discrete variables)
$$ f(y)=\exp \left( \frac{y \theta-b(\theta)}{\phi}w + c(y,\phi,w) \right)$$
where $\theta$ is called the _natural parameter_ (or parameter of interest) and $\phi$ the _dispersion parameter_. 

The Poisson distribution is a discrete distribution with probability mass function
$$ f(y)=\frac{\lambda^{y}}{y!}\exp(- \lambda), \text{ for } y=0,1,\ldots$$
where $\lambda>0$.

**a)** (10 points)

* <span class="question">Show that the Poisson distribution is a univariate exponential family, and specify what the elements of the exponential family ($\theta$, $\phi$, $b(\theta)$, $w$, $c(y,\phi,w)$) are.</span>
* <span class="question">What is the connections between $\text{E}(Y)$ and elements of the exponential family? </span>
* <span class="question">What is the connections between $\text{Var}(Y)$ and elements of the exponential family?</span>
* <span class="question">Use these connections to derive the mean and variance for the Poisson distribution.</span>
* <span class="question">If the Poisson distribution is used as the distribution for the response in a generalized linear model, what is the _canonical link_ function?</span>

**b)** (15 points)

We consider a Poisson regression with log link $\eta_i=g(\mu_i)=\ln(\mu_i)$, and linear predictor equal to $\eta_i={\bf x}_i^T \boldsymbol{\beta}$. Further, let $p$ be the number of regression parameters in $\boldsymbol{\beta}$ (intercept included). The response--covariate pairs $(Y_i, {\bf x}_i)$ are independent for $i=1,\ldots, n$.

* <span class="question">Does this set-up satisfy the _requirements_ of a GLM model? Explain.</span>
* <span class="question">Write down the log-likelihood.</span>
* <span class="question">From the log-likelihood, _derive_ the formula for the score function ${\bf s}(\boldsymbol{\beta})$ and the expected Fisher information matrix, ${\bf F}(\boldsymbol{\beta})$.</span>
* <span class="question">What are the dimensions of ${\bf s}(\boldsymbol{\beta})$ and ${\bf F}(\boldsymbol{\beta})$?</span>
* <span class="question">How can ${\bf s}(\boldsymbol{\beta})$ and ${\bf F}(\boldsymbol{\beta})$ be used to arrive at a maximum likelihood estimate for $\boldsymbol{\beta}$?</span>

**c)** (10 points)

We now look at a data set giving the number of species of tortoise on the various Galapagos Islands (Data taken from the book "Practical Regression and Anova using R" by Julian J. Faraway.).

The data set contains measurements on 30 islands, and we study the following variables:

* `Species`: The number of species of tortoise found on the island.
* `Area`: The area of the island (km$^2$).
* `Elevation`: The highest elevation of the island (m).
* `Nearest`: The distance from the nearest island (km).
* `Scruz`: The distance from Santa Cruz island (km).
* `Adjacent`: The area of the adjacent island (km$^2$).


We have fitted a Poisson regression with log link to `Species` as response, and the other five variables are used as continuous covariates. Print-out from the fitted model is given in below.

```{r}
library(faraway)
fit <- glm(Species~Area+Elevation+Nearest+Scruz+Adjacent,data=gala,family=poisson(link=log))
summary(fit)
```

Let $\boldsymbol{\beta}$ be a $6 \times 1$ column vector with the regression coefficients (intercept included), and let $\hat{\boldsymbol{\beta}}$ be the maximum likelihood estimator for $\boldsymbol{\beta}$. 

* <span class="question">Write down the asymptotic distribution for $\hat{\boldsymbol{\beta}}$, and specify how the covariance matrix for $\hat{\boldsymbol{\beta}}$ is estimated.</span>

We will focus on the effect of `Elevation`, and denote the corresponding regression coefficient $\beta_2$. 

* <span class="question">Write down the maximum likelihood estimate for $\beta_2$ in the print-out above.</span>
* <span class="question">How can you explain this value to a biologist interested in understanding the effect of `Elevation` on the number of species of tortoise found on the islands? </span>
* <span class="question">What is numerical value for the estimated standard deviation of $\hat{\beta}_2$ given in above?</span>
* <span class="question">Construct an approximate 95\% confidence interval for $\beta_2$.</span>

## Problem 3: Exam December 2017 from UiO, Problem 1.
(written out - with small changes to fit the notation we use - from <https://www.uio.no/studier/emner/matnat/math/STK3100/h17/stk3100-4100_2017_2eng.pdf>)

Assume that the random variable $Y$ is Poisson distributed with probability mass function (pmf)

$$\text{P}(Y = y | \lambda)= \frac{\lambda^y}{y!} \exp(-\lambda), \ y = 0, 1, 2, \dots.$$

<span class="question">a) Show that the distribution of $Y$ is an exponential family, that is, show that the pmf can be written in the form
$$\exp\left\{\frac{y \theta - b(\theta)}{\phi}w + c(y, \phi, w)\right\},$$
and determine $\theta$, $\phi$, $w$, $b(\theta)$ and $c(y, \phi, w)$.</span>

We then assume that $Y_1, Y_2, \dots, Y_n$ are independent with the pmf from a), and let $\mu_i = \text{E}(Y_i)$, $i = 1, \dots, n$.

<span class="question">b) Explain what we mean by a generalized linear model (GLM) for $Y_1, Y_2, \dots, Y_n$ with link function $g(\cdot)$, and determine the canonical link function.</span>

<span class="question">c) Derive an expression for the log-likelihood function $l(\mathbf{\mu}; \mathbf{y})$, where $\mathbf{y} = (y_1, \dots, y_n)^T$ is the observed value of $\mathbf{Y} = (Y_1, Y_2, \dots, Y_n)^T$ and $\mathbf{\mu} = (\mu_1, \dots, \mu_n)^T$.</span>

<span class="question">d) Explain what we mean by a saturated model and determine the maximum of $l(\mathbf{\mu}; \mathbf{y})$ for the saturated model.</span>

<span class="question">e) Explain what we mean by the deviance $D(\mathbf{y}; \mathbf{\hat{\mu}})$ of a Poisson GLM, find an expression for the deviance, and discuss how it may be used.</span>



# R packages

```{r, eval=FALSE}
install.packages(c("tidyverse", 
  "ggplot2", 
  "statmod",
  "corrplot", 
  "ggplot2", 
  "GGally",
  "boot"))
```

# Further reading
* A. Agresti (1996): "An Introduction to Categorical Data Analysis".
* A. Agresti (2015): "Foundations of Linear and Generalized Linear Models." Wiley.
* A. J. Dobson and A. G. Barnett (2008): "An Introduction to Generalized Linear Models", Third edition. 
* J. Faraway (2015): "Extending the Linear Model with R", Second Edition. <http://www.maths.bath.ac.uk/~jjf23/ELM/>
* P. McCullagh and J. A. Nelder (1989): "Generalized Linear Models". Second edition.

