---
title: "TMA4315 Generalized linear models H2018"
subtitle: "Module 8: Generalized linear mixed effects models (GLMM)"
author: "Mette Langaas and Ingeborg Hem, Department of Mathematical Sciences, NTNU"
date: "15.11.2018 [PL], 16.11.2018 [IL]"
output: #3rd letter intentation hierarchy
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
#  pdf_document:
  #   toc: true
  #   toc_depth: 2
  #   keep_tex: yes
#  beamer_presentation:
#    keep_tex: yes
#    fig_caption: false
#    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,tidy=TRUE,message=FALSE,warning=FALSE,results="hold")
showsol<-FALSE
```

(Latest changes: 16.11: IL added links, 13.11.2018 - first version).

# Overview

Aim: Present methods for analysing correlated responses in a generalized linear models setting - LMM meets GLM to become _GLMM_.

Also here (as in module LMM) we will only consider _two-level models_ and in particular focus on _random intercept models_ for binomial and Poisson responses. Emphasis will be on understanding.

---


## Learning material

* Textbook: Fahrmeir et al (2013): Chapter 7.5: 389-394, 7.7.
* [Classnotes (15.11.2018)](https://www.math.ntnu.no/emner/TMA4315/2018h/M8PL.pdf)

---

## Topics

* beaches example - revisited
* notation
* the generalized linear mixed effect model (three ingredients)
* the GLMM with random intercept
* the marginal model
* parameter estimation and Laplace approximation
* summing up: what do we need to know about the GLMM?
* [additional info on different software (not on reading list)](#additional)

Jump to [interactive](#interactive).

# Beaches example - revisited
This example is taken from Zuur et al. (2009, chapter 5, pages 101-142), and data are referred to as RIKZ. Data were collected on nine different beaches in the Netherlands
with the aim to investigate if there is a relationship between the 

* richness of species (number of species observed) and 
* NAP: the height of the sampling station compared to mean tidal level.

Data: 45 observations, taken at 9 beaches:

* beach: the beach that the samples were taken, for each beach 5 different samples were taken.

---

```{r,echo=FALSE}
# library("AED")
# data(RIKZ)
# devtools::install_github("https://github.com/romunov/AED")
# write.csv( RIKZ, file="RIKZ.csv")
  
# RIKZ <- read.csv("http://faculty.concordia.ca/pperesne/BIOL_422_680/RIKZ.csv")

RIKZ <- read.csv("RIKZ.csv")

library(lme4)
library(arm)
library(ggplot2)
gg <- ggplot(RIKZ, aes(x = NAP, y = Richness))
gg <- gg + geom_point(color = "blue", alpha = 0.7)
gg <- gg + geom_smooth(method = "lowess", color = "black")
gg <- gg + theme_bw()
gg <- gg + facet_wrap(~Beach)
gg
```

\small

We want to use the richness of species as the response in a regression model.
In Module LMM we assumed that the response could be viewed as a normal variable, but now we rather assume that 
the response comes from a _Poisson distribution_.

**Q**: Why do we want to assume a Poisson distribution?
\normalsize

---

**A**:

We observe events that may occur within a time interval or a region. 

1. The number of events occuring within a time interval or a region, is independent of the number of events that occurs in any other disjoint (non-overlapping) time interval or region.
2. The probability that a single event occurs within a small time interval or region, is proportional to the length of the interval or the size of the region.
3. The probability that more than one event may occur within a small time interval or region is negligable.

When all of these three properties are funfilled we have a _Poisson process_.

* The number of events in a Poisson process follows a Poisson distribution.

---

We go ahead and fit a Poission GLM with log-link and NAP as covariate. In Module LMM we talked about possible solutions to how to handle multiple observations for each beach and came up with:

---

**Possible solutions:**

1. Use all 45 observations in a Poisson regression - with a common intercept and linear effect in NAP.
**Problem:** violation of assumption of independent observations lead to wrong estimates for variances of parameter estimates.

2. Add beach as covariate to Poisson regression - then we estimate one regression coefficient for each beach (intercepts) in addition to the linear effect in NAP.
**Problem:** why do we want to add 8 extra parameters to estimate (why 8?) values for the 9 beaches? Loss of power and what would we use the beach estimates for?

3. Add beach as a _random_ covariate to the Poisson regression: this is called _random intercept models_.
**Problem:** new stuff - slightly complicated. We do this because beaches not of interest in themselves, only random sample from population of beaches, and therefore we only need to account for beaches, not estimate separate parameters.

---

But - first: what do we remember about GLMs (and Poission regression with log-link in particular)?

---

**Assumptions for GLM model (Poisson): **

1. Random component from exponential family: We have for our beaches: $Y_i \sim \text{Poisson}(\lambda_i)$, with $\text{E}(Y_i)=\lambda_i$, and $\text{Var}(Y_i)=\lambda_i$.

2. Linear predictor: $\eta_i={\bf x}_i^T {\boldsymbol \beta}$.

3. Log link
$$\eta_i=\ln(\lambda_i)=g(\lambda_i)$$
and (inverse thereof) response function
$$\lambda_i=\exp(\eta_i)$$
Canonical link for Poisson is log.

Parameter estimation is performed using the method of maximum likelihood, numerically using Fisher scoring.

---

**Solution 1:** all observations in one GLM together - standard errors not correct since assumption of independent obsevations is violated.

Remember: For interpretation. If $x_{i1}$ increases by one unit to $x_{i1}+1$ then the mean $\text{E}(Y_i)$ will in our model change by a _factor_ $\exp(\beta_1)$. Not so easy to compare to the normal case.
\footnotesize
```{r,results="hold"}
fitall=glm(Richness~NAP,data=RIKZ,family=poisson(link=log))
summary(fitall)
fitall$coefficients[2]
exp(fitall$coefficients[2])
```
\normalsize

---

\small
```{r,tidy=FALSE}
library(ggplot2)
library(ggpubr)
df=data.frame("Y"=RIKZ$Richness,"x"=RIKZ$NAP,"fitted"=fitall$fitted.values,"dres"=residuals(fitall,type="deviance"),"pres"=residuals(fitall,type="pearson"),catBeach=as.factor(RIKZ$Beach))
gg1=ggplot(df)+geom_point(aes(x=x,y=Y,colour=catBeach))+geom_line(aes(x=x,y=fitted))+xlab("NAP")+ylab("Richness")
gg2=ggplot(df)+geom_point(aes(x=fitted,y=dres,color=catBeach))+xlab("fitted")+ylab("deviance residuals")
ggarrange(gg1,gg2)  
```

\normalsize
**Q**: What is the formula for the black line on the left panel?
**A**: see classnotes.

---

**Solution 2:** fixed effects for each beach - many estimates not so much of interest when population is in focus.
\footnotesize
```{r}
RIKZ$beachfactor=as.factor(RIKZ$Beach)
fitbeach=glm(Richness~NAP+beachfactor,data=RIKZ,contrasts=list(beachfactor="contr.sum"),family=poisson(link="log"))
summary(fitbeach)
```
\normalsize

---

\small
```{r,tidy=FALSE}
library(ggplot2)
library(ggpubr)
df=data.frame("Y"=RIKZ$Richness,"x"=RIKZ$NAP,"fitted"=fitbeach$fitted.values,"dres"=residuals(fitbeach,type="deviance"),"pres"=residuals(fitbeach,type="pearson"),catBeach=as.factor(RIKZ$Beach))
gg1=ggplot(df)+geom_point(aes(x=x,y=Y,colour=catBeach))+geom_line(aes(x=x,y=fitted))+xlab("NAP")+ylab("Richness")+facet_wrap(~catBeach)
gg2=ggplot(df)+geom_point(aes(x=fitted,y=dres,color=catBeach))+xlab("fitted")+ylab("deviance residuals")
ggarrange(gg1,gg2)  
```
\normalsize

---

# Generalized linear mixed effects models

## Notation 

combining GLM (Module 5) and LMM (Module 7)

We have $n_i$ repeated observations ($j=1,\ldots, n_i$) from each of $i=1,\ldots,m$ clusters (or individuals).

* Responses: $Y_{i1},Y_{i2},\ldots, Y_{in_i}$ (e.g. species richness at beach $i$ sample $j$)
* Covariates: ${\bf x}_{i1},{\bf x}_{i1}, \ldots, {\bf x}_{in_i}$ (e.g. NAP for beach $i$ sample $j$)

The covariates ${\bf x}_{ij}$ are $p\times 1$ vectors (as before, $k$ covariates and one intercept so $p=k+1$).

---

## Repetition GLM

1. **Random component** - exponential family (now $y_{ij}$)

\[ f(y_{ij}\mid \theta_{ij})=\exp \left( \frac{y_{ij} \theta_{ij}-b(\theta_{ij})}{\phi}\cdot w_{ij} + c(y_{ij}, \phi, w_{ij}) \right) \]

* $\theta_{ij}$ is called the canonical parameter and is a parameter of interest

* $\phi$ is called a nuisance parameter (and is not of interest to us=therefore a nuisance (plage))

* $w_{ij}$ is a weight function, in most cases $w_{ij}=1$

* $b$ and $c$ are known functions.

We looked in detail into the binomial, Poisson and gamma distribution (in addition to the normal).

---

2. **Linear predictor** (previously "systematic component"):

$$\eta_{ij}={\bf x}_{ij}^T {\boldsymbol \beta}$$ 

3. **Link function $g$- and response function $h$**

$$\eta_{ij}=g(\mu_{ij})$$

$$\mu_{ij}=h(\eta_{ij})$$

---

## New element

we add a random component (effect) to the _linear predictor_!

2. $$\eta_{ij}={\bf x}_{ij}^T {\boldsymbol \beta}+ {\bf u}^T_{ij} \gamma_i$$ 
and we make the same assumptions as for the LMM

$$ \gamma_i \sim N({\bf 0},{\bf Q})$$
and independent from covariates. Observe: we still assume normality for the random effects!

In addition, the mean $\mu_{ij}=\text{E}(Y_{ij}\mid \gamma_i)$ is now to be seen to be conditional on the random effect $\gamma_i$. This means that the specified response distribution is conditional on the random effect: $f(y_{ij}\mid \gamma_i)$.

---

## GLMM model assumptions

### Distributional assumptions
(previously "1. Random component from exponential family")

* Given the random effects $\gamma_i$ and the covariates ${\bf x}_{ij}$ and ${\bf u}_{ij}$
* the responses $Y_{ij}$ are _conditionally independent_ and
* the _conditional distribution_ $f(y_{ij}\mid \gamma_i)$ belongs to and exponential family.

---

### Structural assumptions
(previously "2. Systematic component" and "3. Link function")

* The _conditional mean_ $\mu_{ij}=\text{E}(Y_{ij}\mid \gamma_i)$ is linked to 
* the _linear predictor_ $\eta_{ij}={\bf x}_{ij}^T {\boldsymbol \beta}+ {\bf u}_{ij} \gamma_i$
* through the link function $$\eta_{ij}=g(\mu_{ij})$$ 
* or equivalently through the response function $$\mu_{ij}=h(\eta_{ij})$$
* where $h^{-1}=g$ (inverse functions).

---

### Distributional assumptions for random effects

* The random effects $\gamma_i$, $i=1,\ldots,m$ are independent and identically distributed
$$ \gamma_i \sim N({\bf 0}, {\bf Q})$$
* where the covariance matrix ${\bf Q}$ is a $(q+1)\times (q+1)$ positive definite.
* An important special case is ${\bf Q}=diag(\tau_0^2, \tau_1^2,\ldots,\tau_q^2)$.

**Q**: Any questions? Notice that we still have normal distribution for the random effects!

**Remark**: for LMMs we used that $\varepsilon \sim N({\bf 0}, \sigma^2 {\bf I})$, but said that it was possible to let the error terms be correlated  $\varepsilon \sim N({\bf 0}, {\bf R})$ where ${\bf R}$ could har non-zero off-diagonal elements. However, for GLMMs this is much more complicated.

---

## Alternative two-step formulation

Consider a random variable $Y_{ij}$ and a random effect $\gamma_i$ for cluster $i$.

Alternative two step formulation:

1. $\gamma_i \sim N({\bf 0},{\bf Q})$
2. Given the realization of $\gamma_i$, then the $Y_{ij}$s are independent and each have the distribution $f(y_{ij}\mid \gamma_i)$ with mean $\mu_{ij}=h(\eta_{ij})$ and linear predictor 
$\eta_{ij}={\bf x}_{ij}^T {\boldsymbol \beta}+ {\bf u}^T_{ij} \gamma_i$.

Remark: this conditional formulation is "easy" to write down, but to do inference we need the marginal distribution of all $Y_{ij}$s together to construct the likelihood. We will soon see that this is hard.

**Q:** 
How would you proceed to generate a GLMM data set - e.g. based on the beaches example?

**A:** see class notes

# GLMM random intercept model

**Distributional assumptions:** we will focus on binomial and Poisson responses.

**Structural assumptions:** The random effect is added to the linear predictor

$$\eta_{ij}={\bf x}_{ij}^T {\boldsymbol \beta}+ \gamma_{0i}$$

and we only consider canonical links, so logit for binomial and log for Poisson.

**Distributional assumptions for random effects**

The $\gamma_{0i}$, $i=1,\ldots,m$ are independent and identically distributed
$$\gamma_{0i}\sim N(0,\tau_0^2)$$

---

## Poisson random intercept model

**Distributional assumptions:** 
The conditional distribution of the response $Y_{ij}$ is Poisson with mean $\lambda_{ij}$
$$
f(y_{ij}\mid \gamma_i)=\frac{\lambda_{ij}^{y_{ij}}}{y_{ij}!}\exp(-\lambda_{ij}) \text{ for } y=0,1,2,\ldots
$$
This is conditional on $\lambda_{ij}$, which means that it is really conditional on the linear predictor - since the mean is a function of the linear predictor, and in the linear predictor we have fixed and random effects. So, conditional on the values for the fixed and random effects.

Then the observations $Y_{ij}$ are _conditionally_ independent for all $i$ and $j$ - but, not _marginally_ independent.

---

**Structural assumptions:**
$$\eta_{ij}={\bf x}_{ij}^T {\boldsymbol \beta}+ \gamma_{0i}$$

$$\ln(\lambda_{ij})=\eta_{ij} \text{ or } \lambda_{ij}=\exp(\eta_{ij})$$

**Distributional assumptions for random effects**

The $\gamma_{0i}$, $i=1,\ldots,m$ are independent and identically distributed
$$\gamma_{0i}\sim N(0,\tau_0^2)$$

**Remark**: It is not necessary that $n_i>1$ to use this model, and this model can be used to take care of overdispersion (when all $n_i=1$).


---

## Beach-example: Poisson GLMM with beach as random intercept 
(we will talk about parameter estimation soon)

```{r,results="hold"}
# library("AED")
# data(RIKZ)
library(lme4)
fitRI=glmer(Richness~NAP +(1|Beach),data=RIKZ,family=poisson(link=log))
summary(fitRI)
```
\normalsize

**Q**: Comment on what you see. 

**A**: `glmer`, "Laplace approximation", we now have $p$-values, and for the fixed effects we read z and not t. No estimate for $\sigma^2$ (or `residual` term in random effects part) - since the model does not contain any dispersion parameter.

---

\tiny

```{r,eval=TRUE}
library(sjPlot)
library(ggplot2)

plot_model(fitRI, type = "re") + ylab("BLUP") + xlab("Beaches")
```

---

```{r}
df <- data.frame(Y = RIKZ$Richness, x = RIKZ$NAP, 
                 fitted = fitted(fitRI),
                 dres = residuals(fitRI,scale=TRUE,type="deviance"),
                 catBeach = as.factor(RIKZ$Beach))
ggplot(df) + 
  geom_point(aes(x = x, y = Y, col = catBeach)) +
  geom_line(aes(x = x, y = fitted)) + 
  xlab("NAP") + ylab("Richness") + 
  facet_wrap(~ catBeach)
```

---

```{r}
xx <- rep(seq(min(RIKZ$NAP), max(RIKZ$NAP), length.out = 100), each = 9)
df <- data.frame(x = xx,
                 eta = coef(fitRI)$Beach[,1] + coef(fitRI)$Beach[,2] * xx,
                 Beach = factor(rep(1:9, times = 100)))
ggplot(df, aes(x = x, y = exp(eta), col = Beach)) + 
  geom_line(size = 1) + 
  labs(x = "NAP", y = "Predicted incidents", 
       title = "Predicted incidents of NAP on Richness") + 
  facet_wrap(~ Beach)
```

---

```{r}
df <- data.frame(fitted = fitted(fitRI), 
                 resid = residuals(fitRI, scaled = TRUE, type = "deviance"),
                 Beach = factor(RIKZ$Beach))
ggplot(df, aes(x = log(fitted), y = resid)) + 
  geom_hline(yintercept = 0) +
  geom_point(aes(col = Beach)) +
  geom_smooth(method = "loess") +
  labs(x = "Log-predicted values", y = "Deviance residuals", title = "Residuals plot")
```

---

```{r}
df <- data.frame(NAP = RIKZ$NAP, 
                 resid = residuals(fitRI, scaled = TRUE,type="deviance"),
                 Beach = factor(RIKZ$Beach))
ggplot(df, aes(x = NAP, y = resid)) + 
  geom_hline(yintercept = 0) +
  geom_point(aes(col = Beach)) +
  geom_smooth(method = "loess") +
  labs(x = "NAP", y = "Residuals", 
       title = "Relationship between predictor and residuals")
```

---

```{r}
df <- data.frame(Beach2 = RIKZ$Beach, 
                 resid = residuals(fitRI, scaled = TRUE,type="deviance"), 
                 Beach = factor(RIKZ$Beach))
ggplot(df, aes(x = Beach2, y = resid)) + 
  geom_hline(yintercept = 0) +
  geom_point(aes(col = Beach)) +
  geom_smooth(method = "loess") +
  labs(x = "Beach", y = "Residuals", 
       title = "Relationship between Beach and residuals")
```

---

```{r}
plot_model(fitRI, type = "diag")$Beach
```

\normalsize

---

## Simulation example 
Aim: to show the difference between random intercept and random slope log Poisson models, and to compare the cluster curves, mean of cluster curves and the marginal mean

This example a slightly modified version (added line and changed to ggplot) of an example in the Lecture notes by Magne Aldrin, Norwegian Computing Centre (which is way the notation differs slightly from what we have used).

---

The model we use is
$$\begin{pmatrix} \gamma_{0i}\\ \gamma_{1i} \end{pmatrix} \sim N \left( \begin{pmatrix}0\\ 0 \end{pmatrix},\begin{pmatrix} 0.2^2 & 0 \\ 0 &0.07^2 \end{pmatrix} \right)$$
and then
$$\eta_{ij} = \beta_0+\beta_1 x_{ij} + \gamma_{0i} $$ for the random intercept model
and 
$$\eta_{ij} = \beta_0+\beta_1 x_{ij} + \gamma_{0i}+\gamma_{1i}x_{ij} $$ for the random slope model.

Values chosen are $\beta_0=-2$ and $\beta_1=0.3$.

Log-link is used so $\eta_{ij}=\ln(\mu_{ij})$. 

Only data for the random intercept and slope are simulated, while parameter estimates are considered known. The number of clusters is $m=1000$ and $n_i = 201$ value of $x$ from $-10$ to $10$ are studied.

---

Below 30 random lines are plotted together (in different colours).
The black dashed line is $\exp(\beta_0+\beta_i x)$ (denoted "not mean" in the plot) and the black solid line is the population mean (really, the average of the $m$ curves).

(Q: what would happen if we simulate data and estimate parameters instead of only plotting means?)

---

\small
```{r, echo=FALSE}
library(MASS)
library(ggplot2)
library(ggpubr)
set.seed(42)


n <- 1000
beta0 <- -2
beta1 <- 0.3
gamma <- mvrnorm(n = n, mu = c(0,0), Sigma = matrix(c(0.2^2, 0, 0, 0.07^2), ncol = 2))
gamma0 <- gamma[,1]; gamma1 <- gamma[,2]
# gamma0 <- rnorm(n = n, mean = 0, sd = 0.2)
# gamma1 <- rnorm(n = n, mean = 0, sd = 0.07)
x <- seq(-10, 10, 0.1)
nx <- length(x)
Mu.ri <- matrix(NA, nrow = nx, ncol = n)
Mu.ris <- matrix(NA, nrow = nx, ncol = n)
for (i in 1:n) {
  eta.ri <- beta0 + gamma0[i] + beta1*x # Random intercept
  Mu.ri[,i] <- exp(eta.ri)
  eta.ris <- beta0 + gamma0[i] + (beta1 + gamma1[i]) * x # Random intercept + slope
  Mu.ris[,i] <- exp(eta.ris)
}
popMu.ri <- rowMeans(Mu.ri)
popMu.ris <- rowMeans(Mu.ris)

df <- data.frame(x = x, y.ri = c(Mu.ri), y.ris = c(Mu.ris),
                 sim = factor(rep(1:n, each = nx)))
df2 <- data.frame(popMu.ri, popMu.ris,
                  notMean = exp(beta0 + beta1*x))

subset <- sort(which(df$sim %in% sample(1:n, 30)))
gg_color_hue <- function(n) {hues <- seq(15, 375, length = n + 1); hcl(h = hues, l = 65, c = 100)[1:n]}
gg1 <- ggplot(data = df[subset,], mapping = aes(x = x, y = y.ri, col = sim)) + geom_line(size = 0.2, show.legend = FALSE) +
  geom_line(data = df2, mapping = aes(x = x, y = popMu.ri, col = "Mean"), inherit.aes = FALSE) +
  geom_line(data = df2, mapping = aes(x = x, y = notMean, col = "Not mean"), inherit.aes = FALSE, linetype = 2) +
  labs(y = "y", title = "Random intercept") + ylim(range(popMu.ri)) +
  guides(colour = guide_legend(override.aes = list(col = "black", linetype = c(1,2)))) + 
  scale_color_manual(values = c(gg_color_hue(30), "black", "black"), breaks = c("Mean", "Not mean"), name = "")
gg2 <- ggplot(data = df[subset,], mapping = aes(x = x, y = y.ris, col = sim)) + geom_line(size = 0.2) + 
  geom_line(data = df2, mapping = aes(x = x, y = popMu.ris, col = "Mean"), inherit.aes = FALSE) +
  geom_line(data = df2, mapping = aes(x = x, y = notMean, col = "Not mean"), inherit.aes = FALSE, linetype = 2) +
  labs(y = "y", title = "Random intercept \nand slope") + 
  xlim(0, max(x)) + ylim(range(popMu.ris)) +
  guides(colour = guide_legend(override.aes = list(col = "black", linetype = c(1,2)))) + 
  scale_color_manual(values = c(gg_color_hue(30), "black", "black"), breaks = c("Mean", "Not mean"), name = "")
ggarrange(gg1, gg2, common.legend = TRUE)

```
\normalsize

**Q**: comment on what you see. Would you expect that the solid and dashed curve were identical?

<!--Observe: The interpretation of the population parameter $\beta$ is not the same in the conditional as in the marginal model (as is the case for the LMM).-->

---


## Binomial random intercept model

We will study this model in the interactive session.
```{r,eval=FALSE,echo=FALSE}
sjp.glmer(fitRI, type = "eff", show.ci = TRUE)
sjp.glmer(fitRI, type = "pred", vars = "NAP")
sjp.glmer(fitRI,
          type = "ri.slope",
          vars = "NAP",
          show.ci = TRUE)
```

### Simulation example
Aim: to show the difference between random intercept and random slope logit models, and to compare the mean cluster curve with the population curve (as for the Poisson model).

This example a slightly modified version (added line and changed to ggplot) of an example in the Lecture notes by Magne Aldrin, Norwegian Computing Centre (which is way the notation differs slightly from what we have used).

---

The model we use is
$$\begin{pmatrix} \gamma_{0i}\\ \gamma_{1i} \end{pmatrix} \sim N \left( \begin{pmatrix}0\\ 0 \end{pmatrix},\begin{pmatrix} 4 & 0 \\ 0 &1 \end{pmatrix} \right)$$
and then
$$\eta_{ij} = \beta_0+\beta_1 x_{ij} + \gamma_{0i} $$ for the random intercept model
and 
$$\eta_{ij} = \beta_0+\beta_1 x_{ij} + \gamma_{0i}+\gamma_{1i}x_{ij} $$ for the random slope model where $\beta_0=3$ and $\beta_1=1$.
Logit-link is used so $\eta_{ij}=\ln(\frac{\pi_{ij}}{1-\pi_{ij}})$. 

Only data for the random intercept and slope are simulated, while parameter estimates are considered known. The number of clusters is $m=1000$ and $n_i = 201$ value of $x$ from $-10$ to $10$ are studied.

---

Below 30 random lines are plotted together (in different colours).
The black dashed line is $\exp(\beta_0+\beta_i x)/(1+\exp(\beta_0+\beta_i x))$ (denoted "not mean" in the plot) and the black solid line is the population mean (really, the average of the $m$ curves).

---

\small
```{r,echo=FALSE}
set.seed(43)
n <- 1000
beta0 <- 3
beta1 <- 1
gamma <- mvrnorm(n = n, mu = c(0,0), Sigma = matrix(c(2^2, 0, 0, 1), ncol = 2))
gamma0 <- gamma[,1]; gamma1 <- gamma[,2]
# gamma0 <- rnorm(n = n, mean = 0, sd = 2)
# gamma1 <- rnorm(n = n, mean = 0, sd = 1)
x <- seq(-10, 10, 0.1)
nx <- length(x)
Mu.ri <- matrix(NA, nrow = nx, ncol = n)
Mu.ris <- matrix(NA, nrow = nx, ncol = n)
for (i in 1:n) {
  eta.ri <- beta0 + gamma0[i] + beta1*x # Random intercept
  Mu.ri[,i] <- exp(eta.ri)/(1 + exp(eta.ri))
  eta.ris <- beta0 + gamma0[i] + (beta1 + gamma1[i]) * x # Random intercept + slope
  Mu.ris[,i] <- exp(eta.ris)/(1 + exp(eta.ris))
}
popMu.ri <- rowMeans(Mu.ri)
popMu.ris <- rowMeans(Mu.ris)

df <- data.frame(x = x, y.ri = c(Mu.ri), y.ris = c(Mu.ris),
                 sim = factor(rep(1:n, each = nx)))
df2 <- data.frame(x = x, popMu.ri, popMu.ris,
                  notMean = exp(beta0 + beta1*x)/(1 + exp(beta0 + beta1*x)))

subset <- sort(which(df$sim %in% sample(1:n, 30)))
gg_color_hue <- function(n) {hues <- seq(15, 375, length = n + 1); hcl(h = hues, l = 65, c = 100)[1:n]}
gg1 <- ggplot(data = df[subset,], mapping = aes(x = x, y = y.ri, col = sim)) + geom_line(size = 0.2) + 
  geom_line(data = df2, mapping = aes(x = x, y = popMu.ri, col = "Mean"), inherit.aes = FALSE) +
  geom_line(data = df2, mapping = aes(x = x, y = notMean, col = "Not mean"), inherit.aes = FALSE, linetype = 2) +
  labs(y = "y", title = "Random intercept") +
  guides(colour = guide_legend(override.aes = list(col = "black", linetype = c(1,2)))) + 
  scale_color_manual(values = c(gg_color_hue(30), "black", "black"), breaks = c("Mean", "Not mean"), name = "")
gg2 <- ggplot(data = df[subset,], mapping = aes(x = x, y = y.ris, col = sim)) + geom_line(size = 0.2) + 
  geom_line(data = df2, mapping = aes(x = x, y = popMu.ri, col = "Mean"), inherit.aes = FALSE) +
  geom_line(data = df2, mapping = aes(x = x, y = notMean, col = "Not mean"), inherit.aes = FALSE, linetype = 2) +
  labs(y = "y", title = "Random intercept \nand slope") +
  guides(colour = guide_legend(override.aes = list(col = "black", linetype = c(1,2)))) + 
  scale_color_manual(values = c(gg_color_hue(30), "black", "black"), breaks = c("Mean", "Not mean"), name = "")
ggarrange(gg1, gg2, common.legend = TRUE)
```

\normalsize

**Q**: comment on what you see.

# Conditional and marginal model

To do parameter estimation in the LMM we started with the marginal model for $Y_{ij}$.

$$f(y_{ij})=\int_{\gamma_i} f(y_{ij}\mid \gamma_i) f(\gamma_i) d\gamma_i$$

where
$$f(\gamma_i) \text{ is } N({\bf 0},{\bf Q})$$
and $f(y_{ij}\mid \gamma_i)$ might be the binomial, Poisson, normal, gamma, ... distribution - chosen as an exponential family.

It turns out that this integral can only be written out analytically in special cases.

In the special case that $f(y_{ij}\mid \gamma_i)$ is normal we saw in Module LMM that the marginal distribution was also multivariate normal.

---

Another special case is the log-linear Poisson random intercept where all $n_i=1$. See pages 393-394 in our textbook to see the solution (that can be used to handle overdispersion in Poisson regression).

Remark: for the random intercept Poisson GLMM with log-link the expected marginal mean is (Agresti, 2015, page 309)

$$
\text{E}(Y_{ij})=\text{E}[\text{E}(Y_{ij}\mid \gamma_{0i})]=\exp({\bf x}_{ij}^T{\boldsymbol \beta}+\tau_0^2/2)
$$

The solid line in the plot to the left above gives this curve, while the dashed is $\exp({\bf x}_{ij}^T{\boldsymbol \beta})$. In Agresti (2015, page 310) the variance of the marginal distribution is given as

$$
\text{Var}(Y_{ij})=\cdots = \text{E}(Y_{ij})+[\text{E}(Y_{ij})^2(\exp(\tau_0^2)-1)]
$$

so we see that the variance is larger than that of a Poisson distribution.

Remark: the conditional mean is $\exp({\bf x}_{ij}^T{\boldsymbol \beta}+\gamma_{0i})$, while the marginal mean is $\exp({\bf x}_{ij}^T{\boldsymbol \beta}+\tau_0^2/2)$.

(If the link function chosen had been the identity link then the marginal mean would have been ${\bf x}_{ij}^T{\boldsymbol \beta}$.)

---
