---
title: "TMA4315 Generalized linear models H2018"
subtitle: "Module 5: Generalized linear models - common core"
author: "Mette Langaas, Department of Mathematical Sciences, NTNU - with contributions from Ingeborg Hem"
date: "11.10.2017 [PL], 12.10.2017 [IL]"
output: #3rd letter intentation hierarchy
#  html_document:
#    toc: true
#    toc_float: true
#    toc_depth: 2
 # pdf_document:
 #   toc: true
 #   toc_depth: 2
 #   keep_tex: yes
  beamer_presentation:
    keep_tex: yes
    fig_caption: false
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,tidy=TRUE,message=FALSE,warning=FALSE,results="hold")
showsol<-TRUE
```

(Latest changes: 11.010, added links to handwritten materials and dispersion formula, 07.10.2018 first version)

# Overview

## Learning material

* Textbook: Fahrmeir et al (2013): Chapter 5.4, 5.8.2.
* [Classnotes 27.09.2018](https://www.math.ntnu.no/emner/TMA4315/2018h/M52018.pdf) 

Additional notes (with theoretical focus):

* [Exponential family from Module 1](https://www.math.ntnu.no/emner/TMA4315/2018h/Module1ExponentialFamily.pdf)
* [Proof of E and Var for exp fam](https://www.math.ntnu.no/emner/TMA4315/2018h/M5ExpFamProofEVar.pdf)
* [Proof of two forms for F](https://www.math.ntnu.no/emner/TMA4315/2018h/ExpectedFisherInfo.pdf)
* [Orthogonal parameters](https://www.math.ntnu.no/emner/TMA4315/2018h/M5OrthPar.pdf) 
* [IRWLS](https://www.math.ntnu.no/emner/TMA4315/2018h/M5irwls.pdf) 

---

## Topics

* random component: exponential family 
    + elements: $\theta$, $\phi$, $w$, $b(\theta)$
    + elements for normal, binomial, Poisson and gamma
    + properties: $\text{E}(Y)=b'(\theta)$ and $\text{Var}(Y)=b''(\theta)\frac{\phi}{w}$ (and proof)
* systematic component= linear predictor
    + requirements: full rank of design matrix
* link function and response function
    + link examples for normal, binomial, Poisson and gamma
    + requirements: one-to-one and twice differentiable
    + canonical link
    
---

* likelihood inference set-up: $\theta_i\leftrightarrow \mu_i\leftrightarrow \eta_i \leftrightarrow \boldsymbol{\beta}$
* the loglikelihood
* the score function
* expected Fisher information matrix for the GLM and covariance for $\hat{\beta}$
    + what about covariance of $\hat{\beta}$ when $\phi$ needs to be estimated?
    + estimator for dispersion parameter
   
* Fisher scoring and iterated reweighted least squares (IRWLS)
* Pearson and deviance statistic
* AIC

-- so, for the first time: no practical examples or data sets to be analysed!

Jump to [interactive](#interactive).

# GLM — three ingredients

## Random component - exponential family
In Module 1 we introduced distributions of the $Y_i$, that could be written in the form of a _univariate exponential family_
$$
f(y_i\mid \theta_i)=\exp \left( \frac{y_i \theta_i-b(\theta_i)}{\phi}\cdot w_i + c(y_i, \phi, w_i) \right)
$$

where we said that 

* $\theta_i$ is called the canonical parameter and is a parameter of interest

* $\phi$ is called a nuisance parameter (and is not of interest to us=therefore a nuisance (plage))

* $w_i$ is a weight function, in most cases $w_i=1$ (NB: can not contain any unknown parameters)

* $b$ and $c$ are known functions.

---


---

### Elements - Poisson

$$
\begin{aligned}
\theta&=\log(\mu) \\
b(\theta) &=e^\theta \\
\phi&=1 \\
w&=1 \\
\text{E}(Y)&=e^\theta \\
\text{Var}(Y)&=\phi/w \\
\end{aligned}
$$
You can get equivalent results for the normal, Bernoulli, and gamma. Here we will look at the general results

---

### Elements - for normal, Bernoulli, Poisson and gamma

We have seen:

```{r, echo = FALSE}
library(knitr)

df <- data.frame(c("normal", "Bernoulli", "Poisson", "gamma"),
                c("$\\mu$", "$\\ln \\left( \\frac{p}{1-p} \\right)$", "$\\ln \\mu$", "$-\\frac{1}{\\mu}$"),
                c("$\\frac{1}{2} \\theta^2$", "$\\ln (1+\\exp(\\theta))$", "$\\exp(\\theta)$", "$-\\ln (-\\theta)$"),
                c("$\\sigma^2$", "$1$", "$1$", "$\\frac{1}{\\nu}$"),
                                c("$1$","$1$",rep("$1$", 2)),
                c("$\\mu=\\theta$","$p=\\frac{\\exp(\\theta)}{1+\\exp(\\theta)}$","$\\lambda=\\exp(\\theta)$","$\\mu=-1/\\theta$"),
                c(rep("$1$", 1),"$p(1-p)$","$\\lambda$","$\\mu^2$"),
                c("$\\sigma^2$","$p(1-p)$","$\\lambda$","$\\mu^2/\\nu$"))
names(df) <- c("Distribution", "$\\theta$", "$b(\\theta)$", "$\\phi$", "$w$", "$\\text{E}(Y)=b'(\\theta)$","$b''(\\theta)$","$\\text{Var}(Y)=b''(\\theta)\\phi/w$")

kable(df)

```

---

## Systematic component - linear predictor

Nothing new - as always in this course: $\eta_i={\bf x}_i^T \boldsymbol{\beta}$, and we require that the $n \times p$ design matrix ${\bf X}=({\bf x}_1^T, {\bf x}_2^T,\ldots, {\bf x}_n^T)$ has full rank (which is $p$). 

Remark: in this course we always assume that $n>>p$.

---

## Link function - and response function

Link function: $\eta_i=g(\mu_i)$

Response function: $\mu_i=h(\eta_i)$

Canonical link: $\eta_i=\theta_i$, so $g(\mu_i)=\theta_i$
<!-- then minimal sufficient statistic exist $\sum_{j=1}^p \beta_j \sum_{i=1}^n y_i x_{ij}$. -->
When the canonical link is used some of the results for the GLM (to be studied in the next sections) are simplified. 

---

### Examples for normal, binomial, Poisson and gamma

```{r, echo = FALSE}

library(knitr)

df <- data.frame(c("normal", "binomial", "Poisson", "gamma"),
                 c("$h(\\eta_i) = \\eta_i$ and $g(\\mu_i) = \\mu_i$, \"identity link\".",
                   "$h(\\eta_i) = \\frac{e^{\\eta_i}}{1+e^{\\eta_i}}$ and $g(\\mu_i) = \\ln \\left(\\frac{\\mu_i}{1-\\mu_i} \\right) = \\text{logit}(p_i)$. NB: $\\mu_i = p_i$ in our set-up.", 
                   "$h(\\eta_i) = \\exp(\\eta_i)$ and $g(\\mu_i) = \\ln(\\mu_i)$, log-link.",
                   "$h(\\eta_i) = -\\frac{1}{\\eta_i}$ and $g(\\mu_i) = -\\frac{1}{\\mu_i}$, negative inverse, or $h(\\eta_i) = \\exp(\\eta_i)$ and $g(\\mu_i) = \\ln(\\mu_i)$, log-ink."))

names(df) <- c("random component", "response function and link function")

kable(df)

```

---

### Requirements

There are a few formal requirements for the mathematics to work, in particular:

* one-to-one (inverse exists)
* twice differential (for score function and expected Fisher information matrix)

---

### Properties of the exponential family

We have two general properties:

$$\text{E}(Y_i)=b'(\theta_i)$$ 
and 

$$\text{Var}(Y_i)=b''(\theta_i)\frac{\phi}{w_i}$$

In class we study the handwritten proof together: [Proof](https://www.math.ntnu.no/emner/TMA4315/2018h/M5ExpFamProofEVar.pdf)

$b''(\theta_i)$ is often called the variance function $v(\mu_i)$.


---

## The Score (as a function of $\theta$)

The score is $\frac{\partial l}{\partial \theta}$, i.e. 

$$
\begin{aligned}
\frac{\partial l_i}{\partial \theta} &= s_i(\theta) = \frac{\partial \left(\frac{y_i \theta_i-b(\theta_i)}{\phi}\cdot w_i + c(y_i, \phi, w_i) \right)}{\partial \theta}
 \\
&= (y_i - b'(\theta))\frac{w_i}{\phi}
\end{aligned}
$$

---

## The Expected Score

As a general result we have $E(s_i(\theta_i)) = 0$

Proof: 

$E(s_i(\theta_i)) = \int \frac{d l(\theta)}{d\theta} f(y|\theta) dy$

and because $d \log(y)/dx = 1/y dy/dx$, we get

$$
E(s_i(\theta_i)) = \int \frac{1}{f(y|\theta)}\frac{d f(y|\theta)}{d\theta} f(y|\theta) dy = \int \frac{d f(y|\theta)}{d\theta} dy
$$

---

Now, if everything is well behaved, we can reverse the integration and differentiation:

$$
E(s_i(\theta_i)) = \int \frac{d (y|\theta)}{d\theta} dy = \frac{d \int (y|\theta) dy }{d\theta} = \frac{d 1 }{d\theta} = 0
$$


---

## A Different Proof that $\text{E}(Y_i)=b'(\theta_i)$

This is straightforward, from $E(s_i(\theta_i)) = 0$

$$
\begin{aligned}
E(s) &= E \left( (y_i - b'(\theta))\frac{w_i}{\phi} \right) \\
&= (E(y_i) - b'(\theta))\frac{w_i}{\phi} =0 \\
&= E(y_i) - b'(\theta) = 0 \\
\end{aligned}
$$
So $E(y_i) = b'(\theta)$

---

# Variance, $Var(Y_i) = b''(\theta)\phi/w$

Strategy: calculate $\partial^2f/\partial \theta^2$, then integrate over $y$

$\int \partial^2f(y)/\partial \theta^2 dy = 0$ (see notes: we can swap integration & partial derivative)

Go to the notes

---

# Observed Fisher Information

The observed Fisher information is 

$$
\begin{aligned}
\frac{\partial^2 l_i}{\partial \theta^2} &= \frac{\partial s_i(\theta)}{\partial \theta} \\
&=\frac{\partial (y_i - b'(\theta))\frac{w_i}{\phi}}{\partial \theta}\\
&=  - b''(\theta)\frac{w_i}{\phi}
\end{aligned}
$$


---


# Likelihood inference set-up

We want to estimate $\boldsymbol{\beta}$, going from $f(Y|\theta)$: 

$\theta_i\leftrightarrow \mu_i\leftrightarrow \eta_i \leftrightarrow \boldsymbol{\beta}$

$$
\begin{aligned}
f(y_i | \theta_i) &= exp \left(\frac{y_i \theta - b(\theta_i)}{\phi/w_i} + c(y_i, \phi, w_i) \right) \\
\theta_i &= b^{'-1}(\mu) (\text{from } \mu_i = b'(\theta_i) (=E(Y_i)))\\\\
 \mu_i &= g^{-1}(\eta_i) \\
\eta_i &= x_i' \boldsymbol{\beta}
\end{aligned}
$$

$b^{'-1}(\mu)$ is horrible. With the canonical link, $\eta_i=\theta_i$, so $g(\mu_i)=\theta_i$.

See class notes or Fahrmeir et al (2015), Section 5.8.2 for the derivation of the loglikelihood, score and expected Fisher information matrix.

---

## Loglikelihood

$$l(\beta)=\sum_{i=1}^n l_i(\beta)=\sum_{i=1}^n\frac{1}{\phi}(y_i\theta_i-b(\theta_i))w_i+\sum_{i=1}^n c(y_i,\phi,w_i)$$

The part of the loglikelihood involving both the data and the parameter of interest is for a _canonical link_ equal to

$$\sum_{i=1}^n y_i \theta_i=\sum_{i=1}^n y_i {\bf x}_i^T\beta=\sum_{i=1}^n y_i \sum_{j=1}^p x_{ij}\beta_j=\sum_{j=1}^p \beta_j \sum_{i=1}^n y_i  x_{ij}$$

---

## Score function

$\theta_i\leftrightarrow \mu_i\leftrightarrow \eta_i \leftrightarrow \boldsymbol{\beta}$

What is the score function as a function of $\beta$? We need a long chain rule...

$$
s(\beta)= \frac{\partial l}{\partial \beta} = \frac{\partial l(\theta)}{\partial \theta} \frac{\partial \theta}{\partial \mu} \frac{\partial \mu}{\partial \eta} \frac{\partial \eta}{\partial \beta}
$$
We already have $\partial l/\partial \theta = (y_i - b'(\theta))\frac{w_i}{\phi}$, so we need the rest

---

## Score function

$$
s(\beta)= \frac{\partial l}{\partial \beta} = \frac{\partial l(\theta)}{\partial \theta} \frac{\partial \theta}{\partial \mu} \frac{\partial \mu}{\partial \eta} \frac{\partial \eta}{\partial \beta}
$$

$$
\begin{aligned}
\frac{\partial l}{\partial \theta_i} &= (y_i - b'(\theta_i))\frac{w_i}{\phi} \\
\frac{\partial \theta_i}{\partial \mu_i} &= \dots \\
\frac{\partial \mu_i}{\partial \eta_i} &= \frac{\partial h(\eta_i)}{\partial \eta_i} = h'(\eta_i) \\
\frac{\partial \eta_i}{\partial \beta} &= \frac{\partial \mathbf{x}_i' \beta}{\partial \beta} = \mathbf{x}_i
\end{aligned}
$$

---

We get $\frac{\partial \theta_i}{\partial \mu_i}$ by reversing numerator and denominator:

$$
\frac{\partial \mu_i}{\partial \theta_i} = \frac{\partial b'(\theta_i)}{\partial \theta_i} = b''(\theta_i) = \frac{w_i \text{Var}(y_i)}{\phi}
$$

So 

$$
\frac{\partial \theta_i}{\partial \mu_i} = \frac{\phi}{w_i \text{Var}(y_i)}
$$

---

# Putting it together

$$
\begin{aligned}
\frac{\partial l}{\partial \theta_i} &= (y_i - b'(\theta_i))\frac{w_i}{\phi} \\
\frac{\partial \theta_i}{\partial \mu_i} &= \frac{\phi}{w_i \text{Var}(y_i)} \\
\frac{\partial \mu_i}{\partial \eta_i} &= \frac{\partial h(\eta_i)}{\partial \eta_i} = h'(\eta_i) \\
\frac{\partial \eta_i}{\partial \beta} &= \frac{\partial \mathbf{x}_i' \beta}{\partial \beta} = \mathbf{x}_i
\end{aligned}
$$
So 

$$
s(\beta) = (y_i - b'(\theta_i))\frac{w_i}{\phi} \frac{\phi}{w_i \text{Var}(y_i)} h'(\eta_i) \mathbf{x}_i =  \frac{(y_i - b'(\theta_i))}{\text{Var}(y_i)} h'(\eta_i) \mathbf{x}_i
$$

---

# Total Score


$$s(\beta)=\sum_{i=1}^n \frac{(y_i-\mu_i){\bf x}_i h'(\eta_i)}{\text{Var}(Y_i)}={\bf X}^T {\bf D} \Sigma^{-1}({\bf y}-{\boldsymbol \mu})$$
where $\Sigma=\text{diag}(\text{Var}(Y_i))$ and ${\bf D}=\text{diag}(h'(\eta_i))$ (derivative wrt $\eta_i$). 

Remark: observe that $s(\beta)=0$ only depends on the distribution of $Y_i$ through $\mu_i$ and $\text{Var}(Y_i)$.

---

# Canonical link

This is neat, because $\frac{\partial \mu_i}{\partial \eta_i}=b''(\theta_i)$:

$$s(\beta)=\sum_{i=1}^n \frac{(y_i-\mu_i){\bf x}_i w_i}{\phi}$$

---

## Expected Fisher information matrix for the GLM and covariance for $\hat{\beta}$

$$F_{[h,l]}(\beta)=\sum_{i=1}^n \frac{x_{ih}x_{il}(h'(\eta_i))^2}{\text{Var}(Y_i)}$$

$$F(\beta)={\bf X}^T {\bf W} {\bf X}$$

where ${\bf W}=\text{diag}(\frac{h'(\eta_i)^2}{\text{Var}(Y_i)})$.

---

Canonical link:

$$\frac{\partial^2 l_i}{\partial \beta_j \partial \beta_l}=- \frac{x_{ij} w_i}{\phi}(\frac{\partial \mu_i}{\partial \beta_l})$$
which do not contain any random variables, so the observed must be equal to the expected Fisher information matrix.


---

## Fisher scoring and iterated reweighted least squares (IRWLS)

Details on the derivation: [IRWLS](https://www.math.ntnu.no/emner/TMA4315/2018h/M5irwls.pdf) 

$$\beta^{(t+1)}=\beta^{(t)} + F(\beta^{(t)})^{-1} s(\beta^{(t)})$$
Insert formulas for expected Fisher information and score function.

$$\beta^{(t+1)}=({\bf X}^T {\bf W}(\beta^{(t)}) {\bf X})^{-1} {\bf X}^T {\bf W}(\beta^{(t)})\tilde{\bf y}_i^{(t)}$$
where ${\bf W}$ is as before ${\bf W}=\text{diag}(\frac{h'(\eta_i)^2}{\text{Var}(Y_i)})$ - but now the current version of ${\boldsymbol{\beta}}^{(t)}$ is used. The diagonal elements are called the _working weights_. The $\tilde{\bf y}_i^{(t)}$ is called the _working response vector_
and has element $i$ given as

$$\tilde{\bf y}_i^{(t)}={\bf x}_i^T\beta^{(t)}+\frac{y_i-h({\bf x}_i^T\beta^{(t)})}{h'({\bf x}_i^T\beta^{(t)})}.$$

Remark: Convergence? With full rank of ${\bf X}$ and positive diagonal elements of ${\bf W}$ we are certain that the inverse will exist, but there might be that the temporary version of ${\bf W}$ can cause problems.

---

See what is output from `glm`- observe working weights as `weights`..
```{r,echo=-1}
library(investr)
fitgrouped=glm(cbind(y, n-y) ~ ldose, family = "binomial", data = investr::beetle) 
#names(fitgrouped)
round(fitgrouped$weights, 2)
round(fitgrouped$residuals, 2)
```

---

## Estimator for dispersion parameter


Let data be grouped as much as possible. With G groups (covariate pattern) with $n_i$ observations for each group (then $n=\sum^G n_i=n$):

$$ \hat{\phi}=\frac{1}{G-p} \sum_{i=1}^G \frac{(y_i-{\hat{\mu}}_i)^2}{b''(\theta_i)/w_i}$$

The motivation behind this estimator is as follows:
$$ \text{Var}(Y_i)=\phi b''(\theta_i)/w_i \Leftrightarrow \phi=\text{Var}(Y_i)/(b''(\theta_i)/w_i)$$


---

# Distribution of the MLE
As before we have that maximum likelihood estimator $\hat{\beta}$ asymptotically follows the multivariate normal distribution with mean $\beta$ and covariance matrix equal to the inverse of the expected Fisher information matrix. This is also true when we replace the unknown $\beta$ with the estimated $\hat{\beta}$ for the expected Fisher information matrix.

$$\hat{\beta} \approx N_p(\beta,F^{-1}(\hat{\beta}))$$
and with 
$$F(\hat{\beta})={\bf X}^T \hat{{\bf W}} {\bf X}$$
where $\hat{{\bf W}}$ denotes that $\hat{\beta}$ is used then calculating ${\bf W}=\text{diag}(\frac{h'(\eta_i)^2}{\text{Var}(Y_i)})$.

---

## What about the distribution of $\hat{\beta}, \hat{\phi}$?

[The concept of orthogonal parameters](https://www.math.ntnu.no/emner/TMA4315/2018h/M5OrthPar.pdf) 


## Hypothesis testing

Same as before - for the Wald we insert the formula for the covariance matrix of $\hat{\beta}$, for the LRT we insert the loglikelihoods and for the score test we insert formulas for the score function and expected Fisher information matrix.

---

# Model assessment and model choice

## Pearson and deviance statistic
Group observations together in groups of maximal size (covariate patterns? interval versions thereof?). Group $i$ has $n_i$ observations, and there are $G$ groups. Asymptotic distribution correct if all groups have big $n_i$. For the non-continuous individual data asymptotic results can not be trusted.

Deviance
$$D=-2[\sum_{i=1}^{g}(l_i(\hat{\mu}_i)-l_i(\bar{y}_i))]$$
with approximate $\chi^2$-distribution with $G-p$ degrees of freedom.

---

Pearson:
$$ X_P^2=\sum_{i=1}^G \frac{(y_i-\hat{\mu}_i)^2}{v(\hat{\mu}_i)/w_i}$$
with approximate $\phi \cdot \chi^2$-distribution with $G-p$ degrees of freedom.

Remember that the variance function $v(\hat{\mu}_i)=b''(\theta_i)$ (this is a function of $\mu_i$ because $\mu_i=b'(\theta_i)$).

## AIC
Let $p$ be the number of regression parameters in our model.
$$\text{AIC} =-2 \cdot l(\hat{\beta})+2p$$
If the dispersion parameter is estimated use $(p+1)$ in place of $p$.


# Further reading

* A. Agresti (2015): "Foundations of Linear and Generalized Linear Models." Wiley.
