---
subtitle: "TMA4315 Generalized linear models H2018"
title: "Module 2: MULTIPLE LINEAR REGRESSION"
author: "Mette Langaas, Department of Mathematical Sciences, NTNU -- with contributions from Øyvind Bakke and Ingeborg Hem"
date: "30.08 and 06.09 [PL], 31.08 and 07.09 [IL]"
output: #3rd letter intentation hierarchy
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
  # pdf_document:
  #  toc: true
  #  toc_depth: 2
#  beamer_presentation:
#    keep_tex: yes
#    fig_caption: false
#    latex_engine: xelatex
---

```{r setup, include=FALSE}
library(formatR)
showsol<-FALSE
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=68),results="hold",tidy=TRUE,warning=FALSE,error=FALSE,message=FALSE)
knitr::opts_chunk$set(echo = TRUE)
```


```{r , echo=FALSE, eval=showsol, warning=FALSE, message=FALSE}
cat("Answers \n")
cat("1.\n")
cat("2.\n")
cat("3.\n")
cat("4.\n")
```

**<a id="secondweek">SECOND WEEK</a>**

* What did we do last week?
* Parameter estimation in practice.
* Statistical inference for parameter estimates 
    + confidence intervals, 
    + prediction intervals, 
    + hypothesis test, 
    + linear hypotheses.
* Introducing deviance - and likelihood ratio test
* Analysis of variance decompositions and $R^2$, sequential ANOVA table.
* Model selection with AIC 

# What to remember from the first week?

### Munich rent index

Munich, 1999: 3082 observations on 9 variables.

* `rent`: the net rent per month (in Euro).
* `rentsqm`: the net rent per month per square meter (in Euro).
* `area`: living area in square meters.
* `yearc`: year of construction.
* `location`: quality of location: a factor indicating whether the location is average location, 1, good location, 2, and top location, 3.
* `bath`: quality of bathroom: a a factor indicating whether the bath facilities are standard, 0, or premium, 1.
* `kitchen`: Quality of kitchen: 0 standard 1 premium.
* `cheating`: central heating: a factor 0 without central heating, 1 with central heating.
* `district`: District in Munich.

More information in Fahrmeir et. al., (2013) page 5.

---

### The GLM way

Independent pairs $(Y_i, {\bf x}_i)$ for $i=1,\ldots,n$.

1. Random component: $Y_i \sim N$ with $\text{E}(Y_i)=\mu_i$ and $\text{Var}(Y_i)=\sigma^2$.
2. Systematic component: $\eta_i={\bf x}_i^T \boldsymbol{\beta}$.
3. Link function: linking the random and systematic component (linear predictor): Identity link and response function.
$\mu_i=\eta_i$.

---

### Likelihood, loglikelihood, score function, observed and expected Fisher information matrix

* Likelihood $L(\beta)=\prod_{i=1}^n f(y_i; \beta)$.
* Loglikelihood $l(\beta)=\ln L(\beta)$.
* Score function $s(\beta)=\frac{\partial l(\beta)}{\partial \beta}$. Find ML estimates by solving $s(\hat{\boldsymbol \beta})={\bf 0}$.
* Observed $H(\boldsymbol{\beta}) = -\frac{\partial^2l(\beta)}{\partial\beta\partial\beta^T}$ and expected Fisher information $F(\beta) =\text{E}(H(\boldsymbol{\beta}))$

---

### Parameter estimators with properties

* Parameter of interest is $\beta$ and $\sigma^2$ is a nuisance. Maximum likelihood estimator 
$$ \hat\beta=({\bf X}^T{\bf X})^{-1} {\bf X}^T {\bf Y}$$
has distribution: $\hat{\beta}\sim N_{p}(\beta,\sigma^2({\bf X}^T{\bf X})^{-1})$.
* Restricted maximum likelihood estimator for ${\bf \sigma}^2$:
\[ \hat{\sigma}^2=\frac{1}{n-p}({\bf Y}-{\bf X}\hat{\beta})^T({\bf Y}-{\bf X}\hat{\beta})=\frac{\text{SSE}}{n-p}\]
with $\frac{(n-p)\hat{\sigma}^2}{\sigma^2} \sim \chi^2_{n-p}$.

---

* Statistic for inference about $\beta_j$, $c_{jj}$ is diagonal element $j$ of $({\bf X}^T{\bf X})^{-1}$.
\[ T_j=\frac{\hat{\beta}_j-\beta_j}{\sqrt{c_{jj}}\hat{\sigma}}\sim t_{n-p}\]
This requires that $\hat{\beta}_j$ and $\hat{\sigma}$ are independent.
* Asymptotically
\[ T_j=\frac{\hat{\beta}_j-\beta_j}{\sqrt{c_{jj}}\hat{\sigma}}\approx N(0,1)\]

Sums of squares of error (SSE): $\text{SSE}=({\bf Y}-{\bf X}{\hat{\beta}})^T({\bf Y}-{\bf X}{\bf \hat{\beta}})=\hat{\varepsilon}^T \hat{\varepsilon}$.

# Parameter estimation in practice

$$ \hat\beta=({\bf X}^T{\bf X})^{-1} {\bf X}^T {\bf Y}$$

**Q**: How is this done in `lm`?

lm(formula, data, subset, weights, na.action,
   method = "qr", model = TRUE, x = FALSE, y = FALSE, qr = TRUE,
   singular.ok = TRUE, contrasts = NULL, offset, ...)

---

## Big data

But, what about big data? Big data are characterized by

* volume
* velocity - data collected in a (near) continuous setting
* variation — many types of data: numerical measurements, images, text
* veracity — quality and thrustworthiness
* value — potential in data?

We need analysis tools that are

* efficent from a computational point of view
* large memory capacity
* can be done automatically
* is sensible from a statistics point of view

---

If the number of observations, $n$, is large a parallel formulation is valuable.

In the simple case where we want to calculate an average, $\hat{\mu}=\sum_{i=1}^n y_i$, we may divide the dataset into $G$ groups and calculate sums (or averages) in each group.
Group sums: $\hat{\mu}_g=\sum_{i:g_i=g} y_i$.
$$ \hat{\mu}=\frac{1}{n}\sum_{i=1}^n y_i=\frac{1}{n}\sum_{g=1}^G \sum_{i:g_i=g} y_i=\frac{1}{n}\sum_{g=1}^G \hat{\mu}_g$$
This makes it possible to calculate the average in parallell operations and put the result together again.

**Q**: Can this also be done for $\hat{\beta}$?

---

### Solutions in R

* `lm` requires memory of order $O(np + p^2)$, which causes problems when $n$ is large.
* The solution `biglm` needs memory of the order $O(p^2)$ where computations are performed in blocks.

Remark: for GLM in general we have no closed form solutions to the $s(\hat{\boldsymbol \beta})={\bf 0}$ so we will use numerical optimization to handle this, and the ´biglm´ also solves the GLM.

# Inference
We will consider confidence intervals and prediction intervals, and then test single and linear hypotheses. Most of this should be known to you from earlier regression courses. We will only focus on the results, and you need to read the details in the derivation by yourself. 

## Confidence intervals (CI)

In addition to providing a parameter estimate for each element of our parameter vector $\beta$ we should also report a $(1-\alpha)100$% confidence interval (CI) for each element. (We will not consider simultanous confidence regions in this course.)

---

We focus on element $j$ of $\beta$, called $\beta_j$. It is known that $T_j =\frac{\hat{\beta}_j-\beta_j}{\sqrt{c_{jj}}\hat{\sigma}}$ follows a $t$-distribution with $n-p$ degrees of freedom. Let $t_{\alpha/2,n-p}$ be such that $P(T_j>t_{\alpha/2,n-p})=\alpha/2$. REMARK: our textbook would here look at area to the left instead of to the right - but we stick with this notation.
Since the $t$-distribution is symmetric around 0, then $P(T_j< -t_{\alpha/2,n-p})=\alpha/2$. We may then write
$$ P(-t_{\alpha/2,n-p}\le T_j \le t_{\alpha/2,n-p})=1-\alpha$$

---

```{r,eval=FALSE,echo=FALSE}
# with old fashion plotting
df=10
f<-function(x) dt(x,df)
par(mar=c(2,1,1,1))
plot(f,-4,4,axes=F,xlab="",ylab="",xaxt="n",type="l")
arrows(-4,0,4.1,0,length=.1)
ymax=1.2*dt(0,df)
arrows(0,-0.1,0,ymax,length=.1)
upper=qt(0.975,df)
lower=-upper
lines(c(lower,lower),c(0,ymax),col="red",lwd=2)
lines(c(upper,upper),c(0,ymax),col="red",lwd=2)
text(lower-0.5,0.04,0.05,cex=1)
text(upper+0.5,0.04,0.05,cex=1)
text(lower,-0.01,expression(-italic(t)[0.05,10]),cex=0.8)
text(upper,-0.01,expression(italic(t)[0.05,10]),cex=0.8)
```

```{r,tidy=FALSE,echo=showsol}
library(ggplot2)
df=10
ggplot(data.frame(x=c(-4,4)), aes(x))+ 
  theme(axis.title.x=element_blank(),axis.title.y=element_blank(),axis.text.x=element_blank(),axis.text.y=element_blank(),axis.ticks = element_blank())+
  stat_function(fun=function(x) dt(x,df), geom="line", colour="red")+
  geom_vline(xintercept=qt(0.975,df),color="blue")+
  geom_vline(xintercept=-qt(0.975,df),color="blue")+
  geom_hline(yintercept=0)+
  annotate("text",x=qt(0.975,df)+0.6,y=0.03,label="alpha/2",parse=TRUE,size=5)+
  annotate("text",x=-qt(0.975,df)-0.6,y=0.03,label="alpha/2",parse=TRUE,size=5)+
  annotate("text",x=0,y=0.03,label="1-alpha",parse=TRUE,size=5)
  
``` 

(Blue lines at $\pm t_{\alpha/2,n-p}$.)

---

Inserting $T_j =\frac{\hat{\beta}_j-\beta_j}{\sqrt{c_{jj}}\hat{\sigma}}$ and solving so $\beta_j$ is in the middle gives:
$$ P(\hat{\beta}_j-t_{\alpha/2,n-p}\sqrt{c_{jj}}\hat{\sigma}
\le \beta_j \le \hat{\beta}_j+t_{\alpha/2,n-p}\sqrt{c_{jj}}\hat{\sigma})=1-\alpha$$

---

A $(1-\alpha)$% CI for $\beta_j$ is when we insert numerical values for the upper and lower limits: $[\hat{\beta}_j-t_{\alpha/2,n-p}\sqrt{c_{jj}}\hat{\sigma},\hat{\beta}_j+t_{\alpha/2,n-p}\sqrt{c_{jj}}\hat{\sigma}]$.

CIs can be found in R using `confint` on an `lm` object. (Here dummy variable coding is used for `location`, with average as reference location.)

\small
```{r}
library(gamlss.data)
fit=lm(rent~area+location+bath+kitchen+cheating,
       data=rent99)
confint(fit)
```
\normalsize

---

**Q** (and A): 

1. What is the interpretation of a 95% confidence interval? 
2. Does the CI for $\hat{\beta}_{\text{area}}$ change if we change the regression model (e.g. not include `cheating`)? 
3. How can we in practice find a CI for `location1` (average location) - when that is not printed above? (Yes, may use formula, but in R without maths?) 
4. What if we go for an asymptotic confidence interval - what will change?

```{r , echo=FALSE, eval=showsol, warning=FALSE, message=FALSE}
cat("Answers \n")
cat("1. See ILw2. \n")
cat("2. Yes! \n")
cat("3. relevel to give good as reference\n")
cat("4. Only quantiles of normal instead of $t$.\n")
```

---

## Prediction intervals

Remember, one aim for regression was to "construct a model to predict the reponse from a set of (one or several) explanatory variables- more or less black box". 

Assume we want to make a prediction (of the response - often called $Y_0$) given specific values for the covariates - often called ${\bf x}_0$. An intuitive point estimate is $\hat{Y}_0={\bf x}_0^T \hat{\beta}$ - but to give a hint of the uncertainty in this prediction we also want to present a prediction interval for the $Y_0$.

---

To arrive at such an estimate we start with the difference between the unobserved response $Y_0$ (for a given covariate vector ${\bf x}_0$) and the point prediction $\hat{Y}_0$, $Y_0-\hat{Y}_0$. 
First, we assume that the unobserved response at covariate ${\bf x}_0$ is independent of our previous observations and follows the same distibution, that is 
$Y_0 \sim N({\bf x}_0^T \beta,\sigma^2)$. Further, 

$$\hat{Y}_0={\bf x}_0^T \hat{\beta} \sim N({\bf x}_0^T \beta,\sigma^2 {\bf x}_0^T({\bf X}^T{\bf X})^{-1}{\bf x}_0).$$

Then, for $Y_0-{\bf x}_0^T \hat{\beta}$ we have
$$\text{E}(Y_0-{\bf x}_0^T \hat{\beta})=0 \text{ and } \text{Var}(Y_0-{\bf x}_0^T \hat{\beta})=\text{Var}(Y_0)+\text{Var}({\bf x}_0^T \hat{\beta})=\sigma^2+\sigma^2{\bf x}_0^T({\bf X}^T{\bf X})^{-1}{\bf x}_0$$
so that 
$$Y_0-{\bf x}_0^T \hat{\beta}\sim N(0,\sigma^2 (1+{\bf x}_0^T({\bf X}^T{\bf X})^{-1}{\bf x}_0)) $$
---

Inserting our REML-estimate for $\sigma^2$ gives 

$$T=\frac{Y_0-{\bf x}_0^T \hat{\beta}}{\hat{\sigma}\sqrt{1+{\bf x}_0^T({\bf X}^T{\bf X})^{-1}{\bf x}_0}}\sim t_{n-p}.$$

Then, we start with 
$$ P(-t_{\alpha/2,n-p}\le \frac{Y_0-{\bf x}_0^T \hat{\beta}}{\hat{\sigma}\sqrt{1+{\bf x}_0^T({\bf X}^T{\bf X})^{-1}{\bf x}_0}} \le t_{\alpha/2,n-p})=1-\alpha$$
and solve so that $Y_0$ is in the middle, which gives
$$P({\bf x}_0^T \hat{\beta}-t_{\alpha/2,n-p}\hat{\sigma}\sqrt{1+{\bf x}_0^T({\bf X}^T{\bf X})^{-1}{\bf x}_0} \le Y_0 \le {\bf x}_0^T \hat{\beta}+t_{\alpha/2,n-p}\hat{\sigma}\sqrt{1+{\bf x}_0^T({\bf X}^T{\bf X})^{-1}{\bf x}_0})=1-\alpha$$

---

A $(1-\alpha)$% PI for $Y_0$ is when we insert numerical values for the upper and lower limits: $[{\bf x}_0^T \hat{\beta}-t_{\alpha/2,n-p}\hat{\sigma}\sqrt{1+{\bf x}_0^T({\bf X}^T{\bf X})^{-1}{\bf x}_0}, {\bf x}_0^T \hat{\beta}+t_{\alpha/2,n-p}\hat{\sigma}\sqrt{1+{\bf x}_0^T({\bf X}^T{\bf X})^{-1}{\bf x}_0}]$.

----

PIs can be found in R using `predict` on an `lm` object, but make sure that `newdata` is a `data.frame` with the same names as the original data. We want to predict the rent - with PI - for an appartment with area 50, location 2 ("good"), nice bath and kitchen and with central heating. 


```{r}
library(gamlss.data)
fit=lm(rent~area+location+bath+kitchen+cheating,
       data=rent99)
newobs=rent99[1,]
newobs[1,]=c(NA,NA,50,NA,2,1,1,1,NA)
predict(fit,newdata=newobs,
        interval="prediction",type="response")
```

---

**Q** (and A): 

1. When is a prediction interval of interest? 
2. Explain the result from `predict` above.
3. What is the interpretation of a 95% prediction
interval? 
4. What will change if want an asymptotic interval?

```{r , echo=FALSE, eval=showsol, warning=FALSE, message=FALSE}
cat("Answers \n")
cat("1. Always? Gives useful information in addition to a point prediction. \n")
cat("2. Fit it point prediction, lwr is lower and upr is upper limit of the 95% PI (default with 95, argument if other.) \n")
cat("3. see ILw2.\n")"
```

---

## Single hypothesis testing set-up

In single hypothesis testing we are interesting in testing one null hypothesis against an alternative hypothesis. In linear regression the hypothesis is often about a regression parameter $\beta_j$:
$$H_0: \beta_j=0 \text{ vs. } H_1: \beta_j\neq 0$$

Remark: we implicitly say that our test is done given that the other variables are present in the model, that is, the other $\beta_i$s ($j\neq i$) are not zero. 

###Two types of errors:

* "Reject $H_0$ when $H_0$ is true"="false positives" = "type I error" ="miscarriage of justice".
These are our _fake news_, which are very important for us to avoid.

* "Fail to reject $H_0$ when $H_1$ is true (and $H_0$ is false)"="false negatives" = "type II error"= "guilty criminal go free".

---

We choose to reject $H_0$ at some significance level $\alpha$ if the $p$-value of the test (see below) is smaller than the chosen significance level.
We say that : Type I error is "controlled" at significance level $\alpha$, which means that the probability of miscarriage of justice (Type I error) does not exceed $\alpha$.

**Q**: Draw a 2 by 2 table showing the connection between 

* "truth" ($H_0$ true or $H_0$ false) - rows in the table, and 
* "action" (reject $H_0$ and accept $H_0$) - columns in the table,

and place the two types of errors in the correct position within the table. 

What else should be written in the last two cells?

---

### Hypothesis test on $\beta_j$ (t-test)

In linear regression models our test statistic for testing $H_0: \beta_j=0$ is 
$$T_0=\frac{\hat{\beta}_j-0}{\sqrt{c_{jj}}\hat{\sigma}}\sim t_{n-p}$$
where $c_{jj}\hat{\sigma}^2=\widehat{\text{Var}}(\hat{\beta}_j)$.

Inserted observed values (and estimates) we have $t_0$.

We would in a two-sided setting reject $H_0$ for large values of $\text{abs}(t_0)$. We may rely on calculating a $p$-value. 

**Q**: what if we want an asymptotic test statistics?

---

### The p-value

A p-value is a test statistic satisfying $0 \leq p({\bf Y}) \leq 1$ for every vector of observations ${\bf Y}$. 

* Small values give evidence that $H_1$ is true. 
* In single hypothesis testing, if the p-value is less than the chosen significance level (chosen upper limit for the probability of committing a type I error), then we reject the null hypothesis, $H_0$. The chosen significance level is often referred to as $\alpha$.
* A p-value is _valid_ if
$$ P(p({\bf Y}) \leq \alpha) \leq \alpha$$
for all $\alpha$, $0 \leq \alpha \leq 1$, whenever $H_0$ is true, that is, if the $p$-value is valid, rejection on the basis of the $p$-value ensures that the probability of type I error does not exceed $\alpha$.
* If $P(p({\bf Y}) \leq \alpha) = \alpha$ for all $\alpha$, $0 \leq \alpha \leq 1$, the $p$-value is called an _exact_ p-value.  

---

In our linear regression we use the $t$-distibution to calculate p-values for our two-sided test situation $H_0: \beta_j=0$ vs. $H_1: \beta_j \neq 0$. Assume we have observed that our test statistic $T_0$ takes the numerical value $t_0$. Since the $t$-distribution is symmetric around $0$ we have

$$p\text{-value}=P(T_0>\text{abs}(t_0))+P(T_0<-\text{abs}(t_0))=2\cdot P(T_0>\text{abs}(t_0)).$$

We reject $H_0$ if our calculated $p$-value is below our chosen signficance level. We often choose as significance level $\alpha=0.05$.

**Q**: what if we want an asymptotic $p$-value?

---

### Munich rent index hypothesis test
We look at print-out using `summary` from fitting `lm`. 

\footnotesize
```{r}
library(gamlss.data)
colnames(rent99)
fit=lm(rent~area+location+bath+kitchen+cheating,
       data=rent99)
summary(fit)
```

\normalsize

---

**Q** (and A):

1. Where is hypothesis testing performed here, and which are the hypotheses rejected at level $0.01$?
2. Will the test statistics and $p$-values change if we change the regression model?
3. What is the relationship between performing an hypothesis test and constructing a CI interval? Remember:
```{r}
library(gamlss.data)
fit=lm(rent~area+location+bath+kitchen+cheating,data=rent99)
confint(fit)
```

```{r , echo=FALSE, eval=showsol, warning=FALSE, message=FALSE}
cat("Answers \n")
cat("1. Column named `t value` gives numerical value for the t-statistic and column `*Pr(>|t|)` gives $p$-value for the two-sided test.\n")
cat("2. Yes. Unless our design matrix has orthogonal columns.\n)" 
cat("3. A: If a $(1-\alpha)$ 100% CI covers the hypothsized value (here 0) then this is a sensible value for the parameter it can be shown that then the $p$-value will be larger than $\alpha$. And, if a $(1-\alpha)$ 100% CI does not cover the hypothsized value (here 0) then this is not a sensible value for the parameter it can be shown that then the $p$-value will be smaller than $\alpha$. \n)"
```

---

## Testing linear hypotheses in regression
We study a normal linear regression model with $p=k+1$ covariates, and refer to this as model A (the larger model). We then want  to investigate the null and alternative hypotheses of the following type(s):
\begin{eqnarray*}
 H_0: \beta_{j}&=&0 \text{ vs. } H_1:\beta_j\neq 0\\
 H_0: \beta_{1}&=&\beta_{2}=\beta_{3}=0 \text{ vs. } H_1:\text{ at least one of these }\neq 0\\
 H_0: \beta_{1}&=&\beta_{2}=\cdots=\beta_{k}=0 \text{ vs. } H_1:\text{ at least one of these }\neq 0\\
 \end{eqnarray*}
 
We call the restricted model (when the null hypotesis is true) model B, or the smaller model.

---

These null hypotheses and alternative hypotheses can all be rewritten as a linear hypothesis 
\[H_0: {\bf C}{\bf \beta}={\bf d} \text{ vs. } {\bf C}{\bf \beta}\neq {\bf d} \]
by specifying ${\bf C}$ to be a $r \times p$ matrix and ${\bf d}$ to be a column vector of length $p$.

The test statistic for performing the test is called $F_{obs}$ and can be formulated in two ways:
\begin{eqnarray}
F_{obs}&=&\frac{\frac{1}{r} (SSE_{H_0}-SSE)}{\frac{SSE}{n-p}} \label{Fobsnested}\\
F_{obs}&=&\frac{1}{r}({\bf C}\hat{{\bf \beta}}-{\bf d})^{\text T}[\hat{\sigma}^2 {\bf C}({\bf X}^{\text T}{\bf X})^{-1}{\bf C}^{\text T}]^{-1}({\bf C}\hat{{\bf \beta}}-{\bf d}) \label{FobsCbeta}
\end{eqnarray}
where $SSE$ is from the larger model A, $SSE_{H_0}$ from the smaller model B, and $\hat{{\bf \beta}}$ and $\hat{\sigma}^2$ are estimators from the larger model A.

----

### Testing a set of parameters - what is ${\bf C}$ and ${\bf d}$?
We consider a regression model with intercept and five covariates, $x_1, \ldots, x_5$. Assume that we want to know if the covariates $x_3$, $x_4$, and $x_5$ can be dropped (due to the fact that none of the corresponding $\beta_j$s are different from zero). This means that we want to test:

$$H_0: \beta_{3}=\beta_{4}=\beta_{5}=0 \text{ vs. } H_1:\text{ at least one of these }\neq 0$$
This means that our ${\bf C}$ is a $6\times 3$ matrix and ${\bf d}$ a $3 \times 1$ column vector
$$ {\bf C}=\begin{pmatrix} 0 &0 &0 &1 &0&0 \\
0&0&0&0&1&0\\
0&0&0&0&0& 1\\
\end{pmatrix} \text{ and } 
{\bf d} =\begin{pmatrix} 0\\0\\0\\ \end{pmatrix}$$

----

### Testing one regression parameter
If we set ${\bf C}=(0,1, 0, \cdots, 0)^T$, a row vector with 1 in position 2 and 0 elsewhere, and ${\bf d}=(0,0,\ldots,0)$, a column vector with 0s, then we test
$$ H_0: \beta_1=0 \text{ vs. } H_1: \beta_1\neq 0.$$
Now ${\bf C}\hat{\beta}=\beta_1$ and ${\bf C}({\bf X}^{\text T}{\bf X})^{-1}{\bf C}^{\text T}=c_{11}$, so that $F_{obs}$ then is equal to the square of the $t$-statistics for testing a single regression parameter.
$$F_{obs}=(\hat{\beta}_1-0)^T[\hat{\sigma}^2 c_{jj}]^{-1}(\hat{\beta}_1-0)=T_1^2$$

Repeat the argument with $\beta_j$ instead of $\beta_1$.

Remark: Remember that $T_{\nu}^2=F_{1,\nu}$.

---

### Testing "significance of the regression"
If we set ${\bf C}=(0,1,1, \cdots ,1)^T$, a row vector with 0 in position 1 and 0 elsewhere, and ${\bf d}=(0,0,\ldots,0)$, a column vector with 0s, then we test
$$ H_0: \beta_1=\beta_2=\cdots= \beta_k =0 \text{ vs. } H_1: \text{at least one different from zero}.$$
This means we test if at least one of the regression parameters (in addition to the intercept) is different from 0. The small model is then the model with only the intercept, and for this model the SSE$_{H_0}$ is equal to SST (sums of squares total, see below). Let SSE be the sums-of-squares of errors for the full model. If we have $k$ regression parameters (in addition to the intercept) then the F-statistic becomes
$$ F_{obs}=\frac{\frac{1}{k}(\text{SST}-\text{SSE})}{\frac{\text{SSE}}{n-p}}$$
with $k$ and $n-p$ degrees of freedom under $H_0$.

---

\footnotesize
```{r}
library(gamlss.data)
fit=lm(rent~area+location+bath+kitchen+cheating,
       data=rent99)
summary(fit)
```
\normalsize

**Q** (and A): Is the regression significant? 
<!-- A: Yes, the F-test is given in the last row of the `summary` print-out and gives a $p$-value clearly below any sensible signficance level. -->

----

### Relation to Wald test

Since $\text{Cov}(\hat{\beta})=\sigma^2 ({\bf X}^T{\bf X})^{-1}$, then $\text{Cov}({\bf C}\hat{\beta})={\bf C}\sigma^2({\bf X}^T{\bf X})^{-1}{\bf C}^T$, so that ${\bf C}\hat{\sigma}^2({\bf X}^T{\bf X})^{-1}{\bf C}^T$ can be seen as an estimate of $\text{Cov}({\bf C}\hat{\beta})$. Therefore, $F_{obs}$ can be written

$$F_{obs}=\frac{1}{r}({\bf C}\hat{{\bf \beta}}-{\bf d})^{\text T}[\widehat{\text{Cov}}({\bf C}\hat{\beta})]^{-1}({\bf C}\hat{{\bf \beta}}-{\bf d})=\frac{1}{r}W$$
where $W$ is a socalled Wald test. It is known that $W\sim \chi^2_r$ asymptotically as $n$ becomes large. We will study the Wald test in more detail later in this course.

---

### Asymptotic result
It can in general be shown that 
$$r F_{r,n-p}\stackrel{n\rightarrow \infty}{\longrightarrow} \chi^2_r.$$
That is, if we have a random variable $F$ that is distributed as Fisher with $r$ (numerator) and $n-p$ (denominator) degrees of freedom, then when $n$ goes to infinity ($p$ kept fixed), then $rF$ is approximately $\chi^2$-distributed with $r$ degrees of freedom.

Also, if our error terms are not normally distributed then we can assume that when the number of observation becomes very large then $rF_{r,n-p}$ is approximately $\chi^2_r$.

# Introducing deviance

The deviance will replace the SSE (sums of squares of errors, aka residual sums of squares) in the GLM setting, and now we take a first look at the deviance, but to do that we first look at the likelihood ratio test.

---

## The likelihood ratio test

An alternative to the Wald test (based on the F-test shown previously) is the likelihood ratio test (LRT), which compares the likelihood of _two models_.

We use the following notation. A: the larger model (this is $H_1$) and B: the smaller model (under $H_0$), and the smaller model is nested within the larger model (that is, B is a submodel of A).

* First we maximize the likelihood for model A (the larger model) and find the maximum likelihood parameter estimates $\hat{\beta}_A$ and $\tilde{\sigma}_A$. The maximum likelihood is achieved at this parameter estimate and is denoted $L(\hat{\beta}_A, \tilde{\sigma}_A)$.
* Then we maximize the likelihood for model B (the smaller model) and find the maximum likelihood parameter estimates $\hat{\beta}_B$ and \tilde{\sigma}_B. The maximum likelihood is achieved at this parameter estimate and is denoted $L(\hat{\beta}_B,\tilde{\sigma}_B)$.

The likelihood of the larger model (A) will always be larger or equal to the likelihood of the smaller model (B). Why? 

---

The likelihood ratio statistic is defined as
$$- 2\ln \lambda=-2(\ln L(\hat{\beta}_B,\tilde{\sigma}_B)-\ln L(\hat{\beta}_A,\tilde{\sigma}_B)) $$ 
(so, $-2$ times small minus large).

Under weak regularity conditions the test statistic is approximately $\chi^2$-distributed with degrees of freedom equal the difference in the number of parameters in the large and the small model. This is general - and not related to the GLM! More about this result in TMA4295 Statistical Inference!

$P$-values are calculated in the upper tail of the $\chi^2$-distribution.

Observe: to perform the test you need to fit both the small and the large model.

Notice: _asymptotically_ the Wald and likelihood ratio test statistics have the same distribution, but the value of the test statistics might be different. 

---

### Example: Munich rent data

* A (larger): model with `area`, `location` and `bath`.
* B (smaller): model with `area` only.

```{r}
library(lmtest)
fitB <- lm(rent ~ area, data = rent99)
fitA <- update(fitB, . ~ . + location+bath)
lrtest(fitB,fitA)

anova(fitB,fitA,test="Chisq")
anova(fitB,fitA)
```

Observe that the LRT can be performed using `anova` with `test="Chisq".

---

## Deviance

The _deviance_ (new!) is used to assess model fit and also for model choice, and is based on the likelihood 
ratio test statistic. It is used for all GLMs in general - and replaces using SSE in multiple linear regression.

First: a covariate pattern is a unique combination of the covariates in our model, for continuous covariates we often have $n$ covariate patterns if we have $n$ observations. Let us assume that for now.

**Saturated model:**
If we were to provide a perfect fit to our data. This means that we have $\hat{\mu}_i=y_i$.
So, each observation is given its own parameter. 

**Candidate model:**
The model that we are investigated can be thought of as a _candidate_ model. Then we maximize the likelihood and get $\hat{\beta}$.

The _deviance_ is then defined as the likelihood ratio statistic, where we put the saturated model in place of the larger model A and our candidate model in place of the smaller model B:

$$D=-2(\ln L(\text{candidate model})-\ln L(\text{saturated model}))$$

The deviance compares the proposed model to the saturated model, and then ask "can we use a more parsimonious model to describe the data as well as the most general model does?".

For the MLR it turns out that the deviance is
$$D=\frac{1}{\sigma^2} \sum_{i=1}^n (y_i-{\bf x}_i^T {\boldsymbol\beta})^2=\frac{\text{SSE}}{\sigma^2}$$

This is sometimes called the _scaled deviance_ while the _unscaled deviance_ is $\phi D$, where $\phi$ is the dispersion parameter. For the normal model the unscaled deviance is thus $\sigma^2 D=\text{SSE}$. 

Warning: both the scaled and unscaled deviance is referred to as the deviance, and called $D$ in different sources. Our textbook use the scaled version, while R use the unscaled. 

# Analysis of variance decomposition and coefficient of determination, $R^2$

It is possible to decompose the total variability in the data, called SST (sums-of-squares total), into a part that is explained by the regression SSR (sums-of-squares regression), and a part that is not explained by the regression SSE (sums-of-squares error, or really residual).

Let $\bar{Y}=\frac{1}{n}\sum_{i=1}^n Y_i$, and $\hat{Y}_i={\bf x}_i^T\hat{\beta}$. Then,
\begin{align*} 
\text{SST}&=\text{SSR}+\text{SSE}\\ 
\text{SST}&=\sum_{i=1}^n (Y_i-\bar{Y})^2={\bf Y}^T({\bf I}-\frac{1}{n}{\bf 1 1}^T){\bf Y}\\
\text{SSR}&=\sum_{i=1}^n (\hat{Y}_i-\bar{Y})^2={\bf Y}^T({\bf H}-\frac{1}{n}{\bf 1 1}^T){\bf Y}\\
\text{SSE}&=\sum_{i=1}^n (Y_i-\hat{Y}_i)^2=\sum_{i=1}^n \hat{\varepsilon}_i^2={\bf Y}^T({\bf I}-{\bf H}){\bf Y}.\\
\end{align*}
The proof can be found in Section 3.5 in our text book Regression.

---

Based on this decomposition we may define the _coefficient of determination_ ($R^2$) as the ratio between SSR and SST, that is 
$$R^2=\text{SSR}/\text{SST}=1-\text{SSE}/\text{SST}$$

1. The interpretation of this coefficient is that the closer it is to 1 the better the fit to the data. If $R^2=1$ then all residuals are zero - that is, perfect fit to the data.
2. In a simple linear regression the $R^2$ equals the squared correlation coefficient between the reponse and the predictor. In multiple linear regression $R^2$ is the squared correlation coefficient between the observed and predicted response.
3. If we have two models $M_1$ and $M_2$, where model $M_2$ is a submodel of model $M_1$, then 
$$ R^2_{M_1}\ge R^2_{M_2}.$$
This can be explained from the fact that $\text{SSE}_{M_1}\le \text{SSE}_{M_2}$.

---

## <a id="seqanova">Analysis of variance tables - with emphasis on sequential Type I ANOVA</a>

It is possible to call the function `anova` on an `lm`-object. What does that function do?
```{r}
library(gamlss.data)
fit=lm(rent~area+location+bath+kitchen+cheating,data=rent99)
anova(fit)
```

----

What is produced is a _sequential_ table of _the reductions in residual sum of squares (SSE) as each term in the regression formula is added in turn_. This type of ANOVA is often referred to as "Type 1" (not to be confused with type I errors). 

We can produce the same table by fitting larger and larger regression models. 

```{r}
library(gamlss.data)
fit=lm(rent~area+location+bath+kitchen+cheating,data=rent99)
fit0 <- lm(rent ~ 1, data = rent99)
fit1 <- update(fit0, . ~ . + area)
fit2 <- update(fit1, . ~ . + location)
fit3 <- update(fit2, . ~ . + bath)
fit4 <- update(fit3, . ~ . + kitchen)
fit5 <- update(fit4, . ~ . + cheating)
anova(fit0, fit1, fit2, fit3, fit4, fit5,test = "F")
#anova(fit0,fit1) # compare model 0 and 1 - NOT sequential
#anova(fit0,fit5) # compare model 0 and 5 - NOT sequential
```

If we had changed the order of adding the covariates to the model, then our anova table would also change. You might check that if you want.

<!-- See the last page of the [classnotes 04.09.2017](https://www.math.ntnu.no/emner/TMA4315/2017h/M2_Classnotes20170904.pdf) for mathematical notation on the sequential test in `anova`, and details on the print-out comes next - NEW: now with formulas! -->

----

###Details on the test `anova(fit)`

When running `anova` on one fitted regression the $F$-test in `anova` is calculated as for "testing linear hypotheses" - but with a slight twist. Our large model is still the full regression model (from the fitted object), but the smaller model is replaced by the _the change from one model to the next_.

Let SSE be the sums-of-squares-error (residual sums of squares) from the full (large, called A) model - this will be our denominator (as always).
For our rent example the denominator will be SSE/(n-p)=64819547/3075 (see print-out above).

However, for the numerator we are not comparing one small model with the full (large) one, we are instead looking at the change in SSE between two (smaller) models (calles model B1 and B2). So, now we have in the numerator the difference in SSE between models B1 and B2, scaled with the difference in number of parameters estimated in model B1 and B2 ="number in B2 minus in B1" (which is the same as the difference in degrees of freedom for the two models).

So, B1 could be the model with only intercept, and B2 could be the model with intercept and area. Then we calculate the SSE for model B1 and for model B2, and keep the difference (here 40299098). Then we count the number of parameters in model B1 and model B2 and compute "number in B2 minus in B1" (here 1). We could istead have calculated the number of degrees of freedom in the smaller model minus the number of degrees of freedom for the larger model (which here is 1). Our numerator is then 40299098/1.

This means that the test statistics we use are:

$$ F_0=\frac{\frac{\text{SSE}_{B1}-\text{SSE}_{B2}}{\text{df}_{B1}-\text{df}_{B2}}}{\frac{\text{SSE}_A}{\text{df}_A}}$$
Remark: notice that the denominator is just the $\hat{\sigma^2}$ from the larger model A.

This makes our $F$-test statistic: $f_0=\frac{40299098/1}{64819547/3075}=1911.765$ (remember that we swap from capital to small letters when we insert numerical values).

To produce a $p$-value to the test that
$$H_0: \text{"Model B1 and B2 are equally good" vs }H_1:\text{"Model B2 is better than B1}$$
and then the $F\sim {\text{df}_{B1}-\text{df}_{B2},\text{df}_A}$. 
In our example we compare to an F-distribution with 1 and 3075 degrees of freedom. The $p$-value is the "probability of observing a test statistic at least as extreme as we have" so we calculate the $p$-value as $P(F>f_0)$. This gives a $p$-value that is practically 0.

If you then want to use the asymptotic version (relating to a chi-square instead of the F), then multiply your F-statistic with $\text{df}_{B1}-\text{df}_{B2}$ and relate to a $\chi^2$ distribution with $\text{df}_{B1}-\text{df}_{B2}$ degrees of freedom, where $\text{df}_{B1}-\text{df}_{B2}$ is the difference in number of parameters in models B1 and B2. In our example $\text{df}_{B1}-\text{df}_{B2}=1$. 


For the anova table we do this sequentially for all models from starting with only intercept to the full model A. This means you need to calculate SSE and df for models of all sizes to calculate lots of these $F_0$s. Assume that we have 4 covariates that are added to the model, and call the 5 possible models (given the order of adding the covariates) 

* model 1: model with only intercept

* model 2: model with intercept and covariate 1

* model 3: model with intercept and covariate 1 and covariate 2

* model 4: model with intercept and covariate 1 and covariate 2 and covariate 3

* model 5: model with intercept and covariate 1 and covariate 2 and covariate 3 and covariate 4

Fit a linear model (lm) for each model 1-5, and store SSE and degrees of freedom=df (number of observations minus number of covariates estimated) for each of the models. Call these SSE$_1$ to SSE$_5$ and df$_1$ to $df$_5$.

The anova output has columns: Df   Sum Sq  Mean Sq  F value    Pr(>F) 
and one row for each covariate added to the model.

* model 2 vs model 1: Df=df$_1$-df$_2$, Sum Sq=SSE$_1$-SSE$_2$, Mean Sq=Sum Sq/Df, F value=(Mean Sq)/(SSE$_5$/df$_5$)=$f_0$, Pr(>F)=pvalue=$P(F>f_0)$.

* model 3 vs model 2: Df=df$_2$-df$_3$, Sum Sq=SSE$_2$-SSE$_3$, Mean Sq=Sum Sq/Df, F value=(Mean Sq)/(SSE$_5$/df$_5$)=$f_0$, Pr(>F)=pvalue=$P(F>f_0)$.

* model 4 vs model 3: Df=df$_3$-df$_4$, Sum Sq=SSE$_3$-SSE$_4$, Mean Sq=Sum Sq/Df, F value=(Mean Sq)/(SSE$_5$/df$_5$)=$f_0$, Pr(>F)=pvalue=$P(F>f_0)$.

* model 5 vs model 4: Df=df$_4$-df$_5$, Sum Sq=SSE$_4$-SSE$_5$, Mean Sq=Sum Sq/Df, F value=(Mean Sq)/(SSE$_5$/df$_5$)=$f_0$, Pr(>F)=pvalue=$P(F>f_0)$.

In R the p-value is calculated as `1-pf(f0,Df)` or as `1-pchisq(Df*f0,Df)` if the asymptotic chisquare distribution is used.

So, this is what is presented - a sequential record of the effect of adding a new covariate.

**Q**: what if you change the order of the covariates into the model? Yes, then everything changes. That is the drawback of Type I (sequential) thinking.

**Q**: What if one of the covariates is a factor? Then all parameters of the factor (e.g. all dummy variables) are tested in one step (more in interactive lecture Problem 2).

---

A competing way of thinking is called _type III ANOVA_ and instead of looking sequentially at adding terms, we (like in `summary`) calculated the contribution to a covariate (or factor) given that all other covariates are present in the regression model. Type III ANOVA is available from library `car` as function `Anova` (possible to give type of anova as input).

**Check** : Take a look at the print-out from  `summary` and  `anova` and observe that for our rent data the $p$-values for each covariate are different due to the different nature of the $H_0$s tested (sequential vs. "all other present").

If we had orthogonal columns for our different covariates the type I and type III ANOVA tables would have been equal.

#### Optional (beyond the scope of this course)

There is also something called a type II ANOVA table, but that is mainly important if we have interactions in our model, so we do not consider that here. If you want to read more this blogplot
<https://www.r-bloggers.com/anova-%E2%80%93-type-iiiiii-ss-explained/>
is a good read. And, in combination with different variants of dummy and effect coding - read this:
<http://rstudio-pubs-static.s3.amazonaws.com/65059_586f394d8eb84f84b1baaf56ffb6b47f.html>. A good read is [Langsrud (2003): ANOVA for unbalanced data: Use Type II instead of Type III sums of squares](https://link.springer.com/article/10.1023/A:1023260610025).

### Pointing ahead

For GLM the sequential analysis of variance (ANOVA) is replace by analysis of deviance.

# Model selection

When we do model selection in the GLM course we will focus on the AIC criterion. The other criteria are added for completeness. We do not use hypothesis tests in model selection.

## Quality measures

To assess the quality of the regression we can report the $R^2$ coefficient of determination. However, since adding covariates to the linear regression can not make the SSE larger, this means that adding covariates can not make the $R^2$ smaller. This means that SSE and $R^2$ are only useful measures for comparing models with the same number of regression parameters estimated.

If we consider two models with the same model complexity then SSE can be used to choose between (or compare) these models.

But, if we want to compare models with different model complexity we need to look at other measures of quality for the regression.


### $R^2$ adjusted (corrected) 
$$R^2_{\text{adj}}=1-\frac{\frac{SSE}{n-p}}{\frac{SST}{n-1}}=1-\frac{n-1}{n-p}(1-R^2)$$
Choose the model with the _largest_ $R^2_{\text{adj}}$.

---

###AIC Akaike information criterion 

AIC is one of the most widely used criteria, and is designed for likelihood-based inference.
Let $l(\hat{\beta}_M,\tilde{\sigma}^2)$ be the maximum of the log-likelihood of the data inserted the maximum likelihood estimates for the regression and nuisance parameter. Further, let $\lvert M \rvert$ be the number of estimated regression parameters (coefficients) in our model, and add 1 if we need to estimate a dispersion parameter (like we do for the normal model).
$$\text{AIC} =-2 \cdot l(\hat{\beta}_M,\tilde{\sigma}^2)+2(\lvert M\rvert +1)$$

For a normal regression model this can be further elaborated on:
$$\text{AIC} =n\ln(\tilde{\sigma}^2)+2(\lvert M\rvert +1)+C$$
where C is a function of $n$ (will be the same for two models for the same data set).
Remark that $\tilde{\sigma}^2=SSE/n$ - our ML estimator (not our unbiased REML), so that the first term in the AIC is just a function of the SSE. For MLR the AIC and the Mallows Cp gives the same result when comparing models.

Choose the model with the minimum AIC. 

---

###BIC  Bayesian information criterion.

The BIC is also based on the likelihood (see notation above).
$$\text{BIC} =-2 \cdot l(\hat{\beta}_M,\tilde{\sigma}^2)+\ln(n)\cdot (\lvert M\rvert +1)$$

For a normal regression model:
$$ \text{BIC}= n\ln(\tilde{\sigma}^2)+\ln(n)(\lvert M\rvert +1)$$
Choose the model with the minimum BIC. 

AIC and BIC are motivated in very different ways, but the final result for the normal regression model is very similar.
BIC has a larger penalty than AIC ($\log(n)$ vs. $2$), and will often give a smaller model (=more parsimonious models) than AIC. 
In general we would not like a model that is too complex.

---

## Model selection strategies

* All subset selection: use smart "leaps and bounds" algorithm, works fine for number of covariates in the order of 40.

* Forward selection: choose starting model (only intercept), then add one new variable at each step - selected to make the best improvement in the model selection criteria. End when no improvement is made.

* Backward elimination: choose starting model (full model), then remove one new variable at each step - selected to make the best improvement in the model selection criteria. End when no improvement is made.

* Stepwise selection: combine forward and backward.

# <a id="interactivew2">Interactive tasks for the second week</a>

### Problem 1: Theory

1. What is the interpretation of a 95% confidence interval? Hint: repeat experiment (on $Y$), on average how many CIs cover the true $\beta_j$? 

<!-- 2. What is the interpretation of a 95% prediction interval? Hint: repeat experiment (on $Y$) for a given ${\bf x}_0$. -->

<!-- 3. Construct a 95% CI for ${\bf x}_0^T \beta$. Explain what is the connections between a CI for $\beta_j$, a CI for ${\bf x}_0^T \beta$ and a PI for $Y$ at ${\bf x}_0$. -->

2. Explain in words and with formulas the $p$-values printed in a `summary` from `lm`.

```{r}
fit=lm(rent~area+location+bath+kitchen+cheating,
       data=rent99)
summary(fit)
```

3. Explain in words and with formulas the full output (with $p$-values) printed in an `anova` from `lm`.

```{r}
anova(fit)
```

4. In particular: why does using `summary` and `anova` on a fitted `lm` gives different $p$-values listed for each covariate. 

Optional: Maybe test out `Anova` in library `car` with type 3 ANOVA to compare?

5. Consider a MLR model $A$ and a submodel $B$ (all parameters in $B$ are in $A$ also). We say that $B$ is nested within $A$. Assume that regression parameters are estimated using maximum likelihood. Why is the following true: the likelihood for model $A$ will always be larger or equal to likelihood for model $B$. 

6. How do we define the deviance of model $A$? What is a _saturated model_ in our MLR setting? What does our finding in 5. imply for the deviance (can the deviance both be positive and negative)?

<!-- 6. Take a look at the details getting from the general AIC criterion to the version used for MLR. We will partly be using AIC for this course, but with likelihood from the binomial, Possion, etc.  -->

---

### Problem 2: Dummy vs. effect coding in MLR (continued)

We have studied the data set with income, place and gender - with focus on dummy variable coding (with different reference levels) and effect coding and the interpretation of parameter estimates. Now we continue with the same data set, but with focus on hypothesis testing (linear hypotheses) and analysis of variance decomposition.

1. Previously, we have read in the data and fitted linear models - look back to see what we found.

```{r}
income <- c(300, 350, 370, 360, 400, 370, 420, 390,
            400,430,420, 410, 300, 320, 310, 305,
            350, 370, 340, 355,370, 380, 360, 365)
gender <- c(rep("Male", 12),rep("Female",12))
place <- rep(c(rep("A",4),rep("B",4),rep("C",4)),2)
data <- data.frame(income,gender=factor(gender,levels=c("Female","Male")),
                   place=factor(place,levels=c("A","B","C")))
```

2. Fit the following model
`model = lm(income~place-1,data=data,x=TRUE)`. Here `x=TRUE` tells the function to calculate the design matrix X, which is stored as `model$x`.

```{r}
model = lm(income~place-1,data=data,x=TRUE)
model$x
```

Examine the results with `summary` and `anova`. What parametrization is used? What is the interpretation of the parameters? Which null hypothesis is tested in the anova-call? What is the result of the hypothesis test?

```{r}
summary(model)
anova(model)
```

3.  Fit the models:
```{r}
model1 = lm(income~place,data=data,x=TRUE,contrasts = list(place="contr.treatment"))
head(model1$x)
summary(model1)
model2 = lm(income ~ place,data=data,x=TRUE,contrasts = list(place="contr.sum"))
head(model2$x)
summary(model2)
```

We have talked about dummy- and effect encoding of categorical covariates. What are the
parametrizations used here? What is the interpretation of the parameters and how do the
parameter interpretations differ between `model1` and `model2`?

4. We want to test the (one-way ANOVA) null hypothesis that there is no effect of place. Use
the $F_{obs}$ to do this both using the dummy-variable and the effect coding of the place-factor. Compare the results from the two coding strategies.

```{r}
model0=lm(income~1,data=data)
anova(model0,model1)
anova(model0,model2)
```


5. Suppose now that there are two factors `place`and `gender`.
```{r}
model3 = lm(income~place+gender,data=data,x=TRUE,contrasts = list(place="contr.treatment",gender="contr.treatment"))
summary(model3)
anova(model3)
model4 = lm(income~place+gender,data=data,x=TRUE,contrasts = list(place="contr.sum",gender="contr.sum"))
summary(model4)
anova(model4)
```
What are the parameterizations? What is the interpretation of the parameters? Does the
ANOVA table look different for the two parametrizations? Hint: orthogonality of design matrix for this balanced design?

6. Finally, fit a model with interactions (model formula is place*gender for both the contrasts and
check if the interaction effect is significant. 

```{r}
model5 = lm(income~place*gender,data=data,x=TRUE,contrasts = list(place="contr.treatment",gender="contr.treatment"))
summary(model5)
anova(model5)
```

```{r}
model6 = lm(income~place*gender,data=data,x=TRUE,contrasts = list(place="contr.sum",gender="contr.sum"))
summary(model6)
anova(model6)
```

---

### Problem 3: Compulsory exercise 1

The TA will give a short introduction to the exercise - and talk about the `lm`-class.

<!-- * Read through how to make an R package together in the group. -->
<!-- * The core is the `mylm` function. What to you have to implement to  -->
<!--    + calculate parameters estimates  -->
<!--    + calculate covariance matrix of the estimated regression coefficients -->
<!--    + perform type 3 hypothesis tests (remember you need to do the asymptotic normal - so no t-distributions) -->
<!--    + what other element to be presented in `summary.lm` must be calculated? $R^2$ -->
<!-- * You also make `summary.mylm`, `print.mylm`,  -->

<!-- MER?    -->

--- 

### Problem 4: Munich Rent index (optional)

Last week all groups decided on using `rent` or `rentsqm` as response, and in short - there was not really a big difference. So, now use `rent` as the response.

1. We now want to use model selection to arrive at a good model. Start by defining which covariates you want to include and how to code them (`location` as dummy or effect coding). What about year of construction - is that a linear covariate? Maybe you want to make intervals in time instead? Linear or categorical for the time? What about the `district`? We leave that since we have not talked about how to use spatial covariates.

Hint: if you want to test out interval versions of year of construction the function `mutate` (from `dplyr`) is useful:
```{r, eval=FALSE}
rent99 <- rent99 %>% mutate(yearc.cat = cut(yearc, breaks = c(-Inf, seq(1920,2000,10)), labels = 10*1:9))
```
More on `dplyr`: Tutorial: <http://genomicsclass.github.io/book/pages/dplyr_tutorial.html>
and Cheat sheet (data wrangling): <https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf> and dplyr in particular: <https://github.com/rstudio/cheatsheets/raw/master/source/pdfs/data-transformation-cheatsheet.pdf>

2. There are many ways to perform model selection in MLR. One possibility is best subsets, which can be done using the `regsubsets` function from library `leaps`. You may define `x` from `model.matrix(fit)[,-1]` (not including the intercept term), and then run `best=regsubsets(x=model.matrix(fit)[,-1],y=rent99$rent)` and look at `summary(best)`.
Explain the print-out (with all the stars). Using the Mallows Cp (named `cp` in the list from `summary(best)`) will give the same result at using AIC (which is not available in this function). What is your preferred model?
Hint: look at the R-code in Problem 2 (Figure 3) from the TMA4267V2017 exam: [pdf](https://www.math.ntnu.no/emner/TMA4267/2017v/Exam/eV2017Enew.pdf), and maybe the solutions for the interprtation [pdf](https://www.math.ntnu.no/emner/TMA4267/2017v/Exam/mergedLFV2017.pdf)

3. Check what is done if you use `stepAIC`. Do you get the same model choice as with best subset selection based on AIC? Why, or why not?

---

### Quiz with Kahoot!

One person on each group go to <https:\\kahoot.it> on a mobile device or a laptop. (The lecturer will hijack the screen for showing questions so you it is difficult to use the PC.)

Give the pin (shown soon) and then give the team nick name "Group1"-"Group8" or make your own personalized group name. Then - if you want - add nicks for all group members. Work together and only provide _one_ answer to each question for each group. In team mode there is a short "team talk" period before you can provide the answer - so you have some time. 1000 points if you answer correctly immediately, 500 if you answer when the time is up, 0 for wrong answers.

# R packages

```{r, eval=FALSE}
install.packages(c("gamlss.data","tidyverse", "GGally", "Matrix","nortest"))
```

# References and further reading

* Slightly different presentation (more focus on multivariate normal theory): [Slides and written material from TMA4267 Linear Statistical Models in 2017, Part 2:  Regression (by Mette Langaas).](https://www.math.ntnu.no/emner/TMA4267/2017v/TMA4267V2017Part2.pdf)
* And, same source, but now [Slides and written material from TMA4267 Linear Statistical Models in 2017, Part 3: Hypothesis testing and ANOVA] (http://www.math.ntnu.no/emner/TMA4267/2017v/TMA4267V2017Part3.pdf)


