---
subtitle: "TMA4315 Generalized linear models H2018"
title: "Module 2: MULTIPLE LINEAR REGRESSION Week 2"
author: "Mette Langaas, Department of Mathematical Sciences, NTNU -- with contributions from Øyvind Bakke and Ingeborg Hem"
date: "30.08.24"
output: #3rd letter intentation hierarchy
 beamer_presentation:
   keep_tex: yes
   fig_caption: false
   latex_engine: xelatex
---

```{r setup, include=FALSE}
showsol<-TRUE
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=68),results="hold",tidy=TRUE,warning=FALSE,error=FALSE,message=FALSE)
knitr::opts_chunk$set(echo = TRUE)
```


## What to remember?

Model: 
$$
{\bf Y}= X {\boldsymbol \beta}+\boldsymbol{\varepsilon}
$$

with full rank design matrix. And classical _normal_ linear regression model when
$$\varepsilon\sim N_n(\bf{0},\sigma^2\bf{I}).$$

Parameter of interest is $\beta$ and $\sigma^2$ is a nuisance. Maximum likelihood estimator 
$$ \hat\beta=({\bf X}^T{\bf X})^{-1} {\bf X}^T {\bf Y}$$
has distribution: $\hat{\beta}\sim N_{p}(\beta,\sigma^2({\bf X}^T{\bf X})^{-1})$.

Restricted maximum likelihood estimator for ${\boldsymbol \sigma}^2$:

$$
\hat{\sigma}^2=\frac{1}{n-p}({\bf Y}-{\bf X}\hat{\beta})^T({\bf Y}-{\bf X}\hat{\beta})=\frac{\text{SSE}}{n-p}
$$

with $\frac{(n-p)\hat{\sigma}^2}{\sigma^2} \sim \chi^2_{n-p}$.

---

Statistic for inference about $\beta_j$, $c_{jj}$ is diagonal element $j$ of $({\bf X}^T{\bf X})^{-1}$.

$$
T_j=\frac{\hat{\beta}_j-\beta_j}{\sqrt{c_{jj}}\hat{\sigma}}\sim t_{n-p}
$$

This requires that $\hat{\beta}_j$ and $\hat{\sigma}$ are independent.

# Inference

We will consider confidence intervals and prediction intervals, and then test single and linear hypotheses.

---

## Confidence intervals (CI)

In addition to providing a parameter estimate for each element of our parameter vector $\beta$ we should also report a $(1-\alpha)100$% confidence interval (CI) for each element. (We will not consider simultanious confidence regions in this course.)

We focus on element $j$ of $\beta$, called $\beta_j$. It is known that $T_j =\frac{\hat{\beta}_j-\beta_j}{\sqrt{c_{jj}}\hat{\sigma}}$ follows a $t$-distribution with $n-p$ degrees of freedom. Let $t_{\alpha/2,n-p}$ be such that $P(T_j>t_{\alpha/2,n-p})=\alpha/2$. 

Since the $t$-distribution is symmetric around 0, then $P(T_j< -t_{\alpha/2,n-p})=\alpha/2$. We may then write
$$P(-t_{\alpha/2,n-p}\le T_j \le t_{\alpha/2,n-p})=1-\alpha$$

---

```{r,eval=FALSE,echo=FALSE}
# with old fashion plotting
df=10
f<-function(x) dt(x,df)
par(mar=c(2,1,1,1))
plot(f,-4,4,axes=F,xlab="",ylab="",xaxt="n",type="l")
arrows(-4,0,4.1,0,length=.1)
ymax=1.2*dt(0,df)
arrows(0,-0.1,0,ymax,length=.1)
upper=qt(0.975,df)
lower=-upper
lines(c(lower,lower),c(0,ymax),col="red",lwd=2)
lines(c(upper,upper),c(0,ymax),col="red",lwd=2)
text(lower-0.5,0.04,0.05,cex=1)
text(upper+0.5,0.04,0.05,cex=1)
text(lower,-0.01,expression(-italic(t)[0.05,10]),cex=0.8)
text(upper,-0.01,expression(italic(t)[0.05,10]),cex=0.8)
```


```{r,tidy=FALSE, echo=FALSE}
library(ggplot2)
df=10
ggplot(data.frame(x=c(-4,4)), aes(x))+ 
  theme(axis.title.x=element_blank(),axis.title.y=element_blank(),axis.text.x=element_blank(),axis.text.y=element_blank(),axis.ticks = element_blank())+
  stat_function(fun=function(x) dt(x,df), geom="line", colour="red")+
  geom_vline(xintercept=qt(0.975,df),color="blue")+
  geom_vline(xintercept=-qt(0.975,df),color="blue")+
  geom_hline(yintercept=0)+
  annotate("text",x=qt(0.975,df)+0.6,y=0.03,label="alpha/2",parse=TRUE,size=5)+
  annotate("text",x=-qt(0.975,df)-0.6,y=0.03,label="alpha/2",parse=TRUE,size=5)+
  annotate("text",x=0,y=0.03,label="1-alpha",parse=TRUE,size=5)
  
``` 

(Blue lines at $\pm t_{\alpha/2,n-p}$.)

----

Inserting $T_j =\frac{\hat{\beta}_j-\beta_j}{\sqrt{c_{jj}}\hat{\sigma}}$ and solving so $\beta_j$ is in the middle gives:

$$ P(\hat{\beta}_j-t_{\alpha/2,n-p}\sqrt{c_{jj}}\hat{\sigma}
\le \beta_j \le \hat{\beta}_j+t_{\alpha/2,n-p}\sqrt{c_{jj}}\hat{\sigma})=1-\alpha$$

A $(1-\alpha)$% CI for $\beta_j$ is when we insert numerical values for the upper and lower limits: $[\hat{\beta}_j-t_{\alpha/2,n-p}\sqrt{c_{jj}}\hat{\sigma},\hat{\beta}_j+t_{\alpha/2,n-p}\sqrt{c_{jj}}\hat{\sigma}]$.

---

CIs can be found in R using `confint` on an `lm` object. (Here dummy variable coding is used for `location`, with average as reference location.)

```{r}
library(gamlss.data)
fit=lm(rent~area+location+bath+kitchen+cheating,data=rent99)
confint(fit)
```

<!---
**Q** (and A): 

1. <span class="question">What is the interpretation of a 95% confidence interval? 
<!-- A: [see Classnotes 04.09.2017](https://www.math.ntnu.no/emner/TMA4315/2017h/M2_Classnotes20170904.pdf)

2. <span class="question">Does the CI for $\hat{\beta}_{\text{area}}$ change if we change the regression model (e.g. not include `cheating`)? 
<!-- A: Yes!

3. <span class="question">How can we in practice find a CI for `location1` (average location) - when that is not printed above? (Yes, may use formula, but in R without maths?) 
<!-- A: relevel to give good as reference. -->

----

## Prediction intervals

Remember, one aim for regression was to "construct a model to predict the reponse from a set of (one or several) explanatory variables- more or less black box". 

Assume we want to make a prediction (of the response - often called $Y_0$) given specific values for the covariates - often called ${\bf x}_0$. An intuitive point estimate is $\hat{Y}_0={\bf x}_0^T \hat{\beta}$ - but to give a hint of the uncertainty in this prediction we also want to present a prediction interval for the $Y_0$.

----

First, we assume that the unobserved response at covariate ${\bf x}_0$ is independent of our previous observations and follows the same distribution, that is 
$Y_0 \sim N({\bf x}_0^T \beta,\sigma^2)$. Further, 

$$\hat{Y}_0={\bf x}_0^T \hat{\beta} \sim N({\bf x}_0^T \beta,\sigma^2 {\bf x}_0^T({\bf X}^T{\bf X})^{-1}{\bf x}_0).$$

Then, for $Y_0-{\bf x}_0^T \hat{\beta}$ we have
$$\text{E}(Y_0-{\bf x}_0^T \hat{\beta})=0 \text{ and } \text{Var}(Y_0-{\bf x}_0^T \hat{\beta})=\text{Var}(Y_0)+\text{Var}({\bf x}_0^T \hat{\beta})=\sigma^2+\sigma^2{\bf x}_0^T({\bf X}^T{\bf X})^{-1}{\bf x}_0$$
so that 
$$Y_0-{\bf x}_0^T \hat{\beta}\sim N(0,\sigma^2 (1+{\bf x}_0^T({\bf X}^T{\bf X})^{-1}{\bf x}_0)) $$
Inserting our REML-estimate for $\sigma^2$ gives 


$$T=\frac{Y_0-{\bf x}_0^T \hat{\beta}}{\hat{\sigma}\sqrt{1+{\bf x}_0^T({\bf X}^T{\bf X})^{-1}{\bf x}_0}}\sim t_{n-p}.$$

---

Then, we start with 
$$ P(-t_{\alpha/2,n-p}\le \frac{Y_0-{\bf x}_0^T \hat{\beta}}{\hat{\sigma}\sqrt{1+{\bf x}_0^T({\bf X}^T{\bf X})^{-1}{\bf x}_0}} \le t_{\alpha/2,n-p})=1-\alpha$$
and solve so that $Y_0$ is in the middle, which gives
$$P({\bf x}_0^T \hat{\beta}-t_{\alpha/2,n-p}\hat{\sigma}\sqrt{1+{\bf x}_0^T({\bf X}^T{\bf X})^{-1}{\bf x}_0} \le Y_0 \le {\bf x}_0^T \hat{\beta}+t_{\alpha/2,n-p}\hat{\sigma}\sqrt{1+{\bf x}_0^T({\bf X}^T{\bf X})^{-1}{\bf x}_0})=1-\alpha$$

A $(1-\alpha)$% PI for $Y_0$ is when we insert numerical values for the upper and lower limits: $[{\bf x}_0^T \hat{\beta}-t_{\alpha/2,n-p}\hat{\sigma}\sqrt{1+{\bf x}_0^T({\bf X}^T{\bf X})^{-1}{\bf x}_0}, {\bf x}_0^T \hat{\beta}+t_{\alpha/2,n-p}\hat{\sigma}\sqrt{1+{\bf x}_0^T({\bf X}^T{\bf X})^{-1}{\bf x}_0}]$.

----

PIs can be found in R using `predict` on an `lm` object, but make sure that `newdata` is a `data.frame` with the same names as the original data. 

We want to predict the rent - with PI - for an apartment with area 50 $m^2$, location 2 ("good"), nice bath and kitchen and with central heating:


```{r}
library(gamlss.data)
fit=lm(rent~area+location+bath+kitchen+cheating,data=rent99)
newobs=rent99[1,]
newobs[1,]=c(NA,NA,50,NA,2,1,1,1,NA)
predict(fit,newdata=newobs,interval="prediction",type="response")
```

--- 

### Questions: 

1. <span class="question">When is a prediction interval of interest? </span>
<!-- A: Always? Gives useful information in addition to a point prediction. -->

2. <span class="question">Explain the result from `predict` above. What are `fit`, `lwr`, `upr`?</span>
<!-- A: Fit it point prediction, lwr is lower and upr is upper limit of the 95% PI (default with 95, argument if other.) -->
3. <span class="question">What is the interpretation of a 95% prediction interval? </span>
<!-- A: [see Classnotes 04.09.2017](https://www.math.ntnu.no/emner/TMA4315/2017h/M2_Classnotes20170904.pdf) -->

---

## Single hypothesis testing set-up

In single hypothesis testing we are interesting in testing one null hypothesis against an alternative hypothesis. In linear regression the hypothesis is often about a regression parameter $\beta_j$:
$$H_0: \beta_j=0 \text{ vs. } H_1: \beta_j\neq 0$$

Remark: we implicitly say that our test is done given that the other variables are present in the model, that is, the other $\beta_i$s ($j\neq i$) are not zero.

---

### Two types of errors:

* "Reject $H_0$ when $H_0$ is true"="false positives" = "type I error" ="miscarriage of justice".
These are our _fake news_, which are very important for us to avoid.

* "Fail to reject $H_0$ when $H_1$ is true (and $H_0$ is false)"="false negatives" = "type II error"= "guilty criminal go free".

---

We choose to reject $H_0$ at some significance level $\alpha$ if the $p$-value of the test (see below) is smaller than the chosen significance level.
We say that : Type I error is "controlled" at significance level $\alpha$, which means that the probability of miscarriage of justice (Type I error) does not exceed $\alpha$.

**Q**: <span class="question">Draw a 2 by 2 table showing the connection between 

* "truth" ($H_0$ true or $H_0$ false) - rows in the table, and 
* "action" (reject $H_0$ and accept $H_0$) - columns in the table,

and place the two types of errors in the correct position within the table. 

What else should be written in the last two cells?

---

### Hypothesis test on $\beta_j$ (t-test)

In linear regression models our test statistic for testing $H_0: \beta_j=0$ is 
$$T_0=\frac{\hat{\beta}_j-0}{\sqrt{c_{jj}}\hat{\sigma}_{\varepsilon}}\sim t_{n-2}$$
where $c_{jj}\hat{\sigma}_{\varepsilon}^2=\widehat{\text{Var}}(\hat{\beta}_j)$.

Inserted observed values (and estimates) we have $t_0$.

We would in a two-sided setting reject $H_0$ for large values of $\text{abs}(t_0)$. We may rely on calculating a $p$-value. 

---

### The p-value

A p-value is a test statistic satisfying $0 \leq p({\bf Y}) \leq 1$ for every vector of observations ${\bf Y}$. 

* Small values are interpreted as evidence that $H_1$ is true(-ish).

* In single hypothesis testing, if the p-value is less than the chosen significance level (chosen upper limit for the probability of committing a type I error), then we reject the null hypothesis, $H_0$. The chosen significance level is often referred to as $\alpha$.

* A p-value is _valid_ if
$$ P(p({\bf Y}) \leq \alpha) \leq \alpha$$
for all $\alpha$, $0 \leq \alpha \leq 1$, whenever $H_0$ is true, that is, if the $p$-value is valid, rejection on the basis of the $p$-value ensures that the probability of type I error does not exceed $\alpha$.

* If $P(p({\bf Y}) \leq \alpha) = \alpha$ for all $\alpha$, $0 \leq \alpha \leq 1$, the $p$-value is called an _exact_ p-value.  

---

In our linear regression we use the $t$-distibution to calculate p-values for our two-sided test situation $H_0: \beta_j=0$ vs. $H_1: \beta_j \neq 0$. Assume we have observed that our test statistic $T_0$ takes the numerical value $t_0$. Since the $t$-distribution is symmetric around $0$ we have

$$p\text{-value}=P(T_0>\text{abs}(t_0))+P(T_0<-\text{abs}(t_0))=2\cdot P(T_0>\text{abs}(t_0)).$$

We reject $H_0$ if our calculated $p$-value is below our chosen signficance level. We often choose as significance level $\alpha=0.05$.

---

### Munich rent index hypothesis test
We look at print-out using `summary` from fitting `lm`. 
```{r}
library(gamlss.data)
fit=lm(rent~area+location+bath+kitchen+cheating,data=rent99)
knitr::kable(summary(fit)$coefficients, digits=3)
```

--- 

**Q** (and A):

1. <span class="question">Where is hypothesis testing performed here, and which are the hypotheses rejected at level $0.01$?
<!-- A: Column named `t value` gives numerical value for the t-statistic and column `*Pr(>|t|)` gives $p$-value for the two-sided test. -->

2. <span class="question">Will the test statistics and $p$-values change if we change the regression model?
<!-- A: Yes. Unless our design matrix has orthogonal columns. -->

3. <span class="question">What is the relationship between performing an hypothesis test and constructing a CI interval? Remember:
```{r}
library(gamlss.data)
fit=lm(rent~area+location+bath+kitchen+cheating,data=rent99)
confint(fit)
```

<!-- A: If a $(1-\alpha)$ 100% CI covers the hypothsized value (here 0) then this is a sensible value for the parameter it can be shown that then the $p$-value will be larger than $\alpha$. And, if a $(1-\alpha)$ 100% CI does not cover the hypothsized value (here 0) then this is not a sensible value for the parameter it can be shown that then the $p$-value will be smaller than $\alpha$. -->

---

## Testing linear hypotheses in regression
We study a normal linear regression model with $p=k+1$ covariates, and refer to this as model A (the larger model). We then want  to investigate the null and alternative hypotheses of the following type(s):
\begin{eqnarray*}
 H_0: \beta_{j}&=&0 \text{ vs. } H_1:\beta_j\neq 0\\
 H_0: \beta_{1}&=&\beta_{2}=\beta_{3}=0 \text{ vs. } H_1:\text{ at least one of these }\neq 0\\
 H_0: \beta_{1}&=&\beta_{2}=\cdots=\beta_{k}=0 \text{ vs. } H_1:\text{ at least one of these }\neq 0\\
 \end{eqnarray*}
We call the restricted model (when the null hypotesis is true) model B, or the smaller model.

---

These null hypotheses and alternative hypotheses can all be rewritten as a linear hypothesis 

$$
H_0: {\bf C}{\boldsymbol \beta}={\bf d} \text{ vs. } {\bf C}{\boldsymbol \beta}\neq {\bf d}
$$

by specifying ${\bf C}$ to be a $r \times p$ matrix and ${\bf d}$ to be a column vector of length $p$.

The test statistic for performing the test is called $F_{obs}$ and can be formulated in two ways:
\begin{eqnarray}
F_{obs}&=&\frac{\frac{1}{r} (SSE_{H_0}-SSE)}{\frac{SSE}{n-p}} \label{Fobsnested}\\
F_{obs}&=&\frac{1}{r}({\bf C}\hat{{\boldsymbol \beta}}-{\bf d})^{\text T}[\hat{\sigma}^2 {\bf C}({\bf X}^{\text T}{\bf X})^{-1}{\bf C}^{\text T}]^{-1}({\bf C}\hat{{\boldsymbol \beta}}-{\bf d}) \label{FobsCbeta}
\end{eqnarray}
where $SSE$ is from the larger model A, $SSE_{H_0}$ from the smaller model B, and $\hat{{\boldsymbol \beta}}$ and $\hat{\sigma}^2$ are estimators from the larger model A.

----

### Testing a set of parameters - what is ${\bf C}$ and ${\bf d}$?
We consider a regression model with intercept and five covariates, $x_1, \ldots, x_5$. Assume that we want to know if the covariates $x_3$, $x_4$, and $x_5$ can be dropped (due to the fact that none of the corresponding $\beta_j$s are different from zero). This means that we want to test:

$$H_0: \beta_{3}=\beta_{4}=\beta_{5}=0 \text{ vs. } H_1:\text{ at least one of these }\neq 0$$
This means that our ${\bf C}$ is a $6\times 3$ matrix and ${\bf d}$ a $3 \times 1$ column vector
$$ {\bf C}=\begin{pmatrix} 0 &0 &0 &1 &0&0 \\
0&0&0&0&1&0\\
0&0&0&0&0& 1\\
\end{pmatrix} \text{ and } 
{\bf d} \begin{pmatrix} 0\\0\\0\\ \end{pmatrix}$$

----

### Testing one regression parameter

If we set ${\bf C}=(0,1, 0, \cdots, 0)^T$, a row vector with 1 in position 2 and 0 elsewhere, and ${\bf d}=(0,0,\ldots,0)$, a column vector with 0s, then we test
$$ H_0: \beta_1=0 \text{ vs. } H_1: \beta_1\neq 0.$$
Now ${\bf C}\hat{\beta}=\beta_1$ and ${\bf C}({\bf X}^{\text T}{\bf X})^{-1}{\bf C}^{\text T}=c_{11}$, so that $F_{obs}$ then is equal to the square of the $t$-statistics for testing a single regression parameter.
$$F_{obs}=(\hat{\beta}_1-0)^T[\hat{\sigma}^2 c_{jj}]^{-1}(\hat{\beta}_1-0)=T_1^2$$

Repeat the argument with $\beta_j$ instead of $\beta_1$.

Remark: Remember that $T_{\nu}^2=F_{1,\nu}$.

---


### Testing "significance of the regression"

If we set ${\bf C}=(0,1,1, \cdots ,1)^T$, a row vector with 0 in position 1 and 0 elsewhere, and ${\bf d}=(0,0,\ldots,0)$, a column vector with 0s, then we test
$$ H_0: \beta_1=\beta_2=\cdots= \beta_k =0 \text{ vs. } H_1: \text{at least one different from zero}.$$
This means we test if at least one of the regression parameters (in addition to the intercept) is different from 0. The small model is then the model with only the intercept, and for this model the SSE$_{H_0}$ is equal to SST (sums of squares total, see below). Let SSE be the sums-of-squares of errors for the full model. If we have $k$ regression parameters (in addition to the intercept) then the F-statistic becomes
$$ F_{obs}=\frac{\frac{1}{k}(\text{SST}-\text{SSE})}{\frac{\text{SSE}}{n-p}}$$
with $k$ and $n-p$ degrees of freedom under $H_0$.

---

Is the regression significant? 

```{r}
summary(fit)$fstatistic

```

<!-- A: Yes, the F-test is given in the last row of the `summary` print-out and gives a $p$-value clearly below any sensible signficance level. -->

----

### Relation to Wald test

Since $\text{Cov}(\hat{\beta})=\sigma^2 ({\bf X}^T{\bf X})^{-1}$, then $\text{Cov}({\bf C}\hat{\beta})={\bf C}\sigma^2({\bf X}^T{\bf X})^{-1}{\bf C}^T$, so that ${\bf C}\hat{\sigma}^2({\bf X}^T{\bf X})^{-1}{\bf C}^T$ can be seen as an estimate of $\text{Cov}({\bf C}\hat{\beta})$. Therefore, $F_{obs}$ can be written

$$F_{obs}=\frac{1}{r}({\bf C}\hat{{\bf \beta}}-{\bf d})^{\text T}[\widehat{\text{Cov}}({\bf C}\hat{\beta})]^{-1}({\bf C}\hat{{\bf \beta}}-{\bf d})=\frac{1}{r}W$$
where $W$ is a so-called Wald test. It is known that $W\sim \chi^2_r$ asymptotically as $n$ becomes large. We will study the Wald test in more detail later in this course.

---

### Asympotic result

It can in general be shown that 
$$r F_{r,n-p}\stackrel{n\rightarrow \infty}{\longrightarrow} \chi^2_r.$$
That is, if we have a random variable $F$ that is distributed as Fisher with $r$ (numerator) and $n-p$ (denominator) degrees of freedom, then when $n$ goes to infinity ($p$ kept fixed), then $rF$ is approximately $\chi^2$-distributed with $r$ degrees of freedom.

Also, if our error terms are not normally distributed then we can assume that when the number of observation becomes very large then $rF_{r,n-p}$ is approximately $\chi^2_r$.

# Focus on likelihood: Likelihood ratio test and deviance

## The likelihood ratio test
An alternative to the Wald test is the likelihood ratio test (LRT), which compares the likelihood of _two models_.

We use the following notation. A: the larger model and B: the smaller model (under $H_0$), and the smaller model is nested within the larger model (that is, B is a submodel of A).

---

* First we maximize the likelihood for model A (the larger model) and find the parameter estimate $\hat{\beta}_A$. The maximum likelihood is achieved at this parameter estimate and is denoted $L(\hat{\beta}_A)$.
* Then we maximize the likelihood for model B (the smaller model) and find the parameter estimate $\hat{\beta}_B$. The maximum likelihood is achieved at this parameter estimate and is denoted $L(\hat{\beta}_B)$.

The likelihood of the larger model (A) will always be larger or equal to the likelihood of the smaller model (B). Why? 

---

The likelihood ratio statistic is defined as
$$- 2\ln \lambda=-2(\ln L(\hat{\beta}_B)-\ln L(\hat{\beta}_A)) $$ 
(so, $-2$ times small minus large).

Under weak regularity conditions the test statistic is approximately $\chi^2$-distributed with degrees of freedom equal the difference in the number of parameters in the large and the small model. This is general - and not related to the GLM! More in TMA4295 Statistical Inference!

$P$-values are calculated in the upper tail of the $\chi^2$-distribution.

Observe: to perform the test you need to fit both the small and the large model.

---

Notice: asymptotically the Wald and likelihood ratio test statistics have the same distribution, but the value of the test statistics might be different. 

```{r}
#TRY OUT BOTH Wald and LRT
```

The LRT can be performed using `anova()`.

---

## Deviance (something new!)

The _deviance_ is used to assess model fit and also for model choice, and is based on the likelihood 
ratio test statistic. It is used for all GLMs in general - and replaces using SSE in multiple linear regression.

**Saturated model:**
If we were to provide a perfect fit to our data t

This "imaginary model" is called the _saturated_ model.
This would be a model where each observation was given its own parameter. 

**Candidate model:**
The model that we are investigated can be thought of as a _candidate_ model. Then we maximize the likelihood and get $\hat{\beta}$.

---

The _deviance_ is then defined as the likelihood ratio statistic, where we put the saturated model in place of the larger model A and our candidate model in place of the smaller model B:

$$D=-2(\ln L(\text{candidate model})-\ln L(\text{saturated model}))$$

For the maximal model, we have parameters $\theta_1, \dots, \theta_n$, where $\theta_i = E[Y_i]$ The log-likelihood for this model is
$$
l(\hat{\beta},{\sigma^2})=-\frac{n}{2}\text{ln}(2\pi)-\frac{n}{2}\text{ln}\sigma^2-\frac{1}{2\sigma^2} ({\bf y}-{\boldsymbol \theta})^T({\bf y}-{\boldsymbol \theta}) = -\frac{n}{2}\text{ln}(2\pi)-\frac{n}{2}\text{ln}\sigma^2
$$

(because $\hat{\theta_i}=y_i$, something you may work out for yourselves)

---

The deviance is then

$$
\begin{aligned}
D &= -2\left(-\frac{n}{2}\text{ln}(2\pi\sigma^2)-\frac{1}{2\sigma^2} ({\bf y}-{\bf X}\hat{\beta})^T({\bf y}-{\bf X}\hat{\beta}) − (-\frac{n}{2}\text{ln}(2\pi\sigma^2)\right) \\
  &= \frac{1}{\sigma^2} ({\bf y}-{\bf X}\hat{\beta})^T({\bf y}-{\bf X}\hat{\beta})
\end{aligned}
$$

Note the connection with the RSS! Under the null hypothesis that the model fits the data
well, $D\sim \chi^2_{n-p}$ exactly (in this case). 


---

# Analysis of variance decomposition and coefficient of determination, $R^2$

It is possible to decompose the total variability in the data, called SST (sums-of-squares total), into a part that is explained by the regression SSR (sums-of-squares regression), and a part that is not explained by the regression SSE (sums-of-squares error, or really residual).

Let $\bar{Y}=\frac{1}{n}\sum_{i=1}^n Y_i$, and $\hat{Y}_i={\bf x}_i^T\hat{\beta}$. Then,
\begin{align*} 
\text{SST}&=\text{SSR}+\text{SSE}\\
\text{SST}&=\sum_{i=1}^n (Y_i-\bar{Y})^2={\bf Y}^T({\bf I}-\frac{1}{n}{\bf 1 1}^T){\bf Y}\\
\text{SSR}&=\sum_{i=1}^n (\hat{Y}_i-\bar{Y})^2={\bf Y}^T({\bf H}-\frac{1}{n}{\bf 1 1}^T){\bf Y}\\
\text{SSE}&=\sum_{i=1}^n (Y_i-\hat{Y}_i)^2=\sum_{i=1}^n \hat{\varepsilon}_i^2={\bf Y}^T({\bf I}-{\bf H}){\bf Y}.\\
\end{align*}

---

Based on this decomposition we may define the _coefficient of determination_ ($R^2$) as the ratio between SSR and SST, that is 
$$R^2=\text{SSR}/\text{SST}=1-\text{SSE}/\text{SST}$$
1. The interpretation of this coefficient is that the closer it is to 1 the better the fit to the data. If $R^2=1$ then all residuals are zero - that is, perfect fit to the data.

2. In a simple linear regression the $R^2$ equals the squared correlation coefficient between the reponse and the predictor. In multiple linear regression $R^2$ is the squared correlation coefficient between the observed and predicted response.

3. If we have two models M1 and M2, where model M2 is a submodel of model M1, then 
$$ R^2_{M_1}\ge R^2_{M_2}.$$
This can be explained from the fact that $\text{SSE}_{M_1}\le \text{SSE}_{M_2}$.
(More in the Theoretical questions.)

----

## <a id="seqanova">Analysis of variance tables - with emphasis on sequential Type I ANOVA</a>

It is possible to call the function `anova` on an `lm`-object. What does that function do?
```{r}
library(gamlss.data)
fit1=lm(rent~area+location+bath,data=rent99)
anova(fit1)
```

----

What is produced is a _sequential_ table of _the reductions in residual sum of squares (SSE) as each term in the regression formula is added in turn_. This type of ANOVA is often referred to as "Type 1" (not to be confused with type I errors). 

We can produce the same table by fitting larger and larger regression models. 

```{r}
library(gamlss.data)
fit=lm(rent~area+location+bath+kitchen+cheating,data=rent99)
fit0 <- lm(rent ~ 1, data = rent99)
fit1 <- update(fit0, . ~ . + area)
fit2 <- update(fit1, . ~ . + location)
fit3 <- update(fit2, . ~ . + bath)

```
---

```{r}
anova(fit0, fit1, fit2, fit3,  test = "F")
#anova(fit0,fit1) # compare model 0 and 1 - NOT sequential
#anova(fit0,fit5) # compare model 0 and 5 - NOT sequential
```

---

If we had changed the order of adding the covariates to the model, then our anova table might also change. You might check that if you want.

---

See the last page of the [classnotes 04.09.2017](https://www.math.ntnu.no/emner/TMA4315/2017h/M2_Classnotes20170904.pdf) for mathematical notation on the sequential test in `anova`, and details on the print-out comes next - NEW: now with formulas!

----

### Details on the test `anova(fit)`

When running `anova` on one fitted regression the $F$-test in `anova` is calculated as for "testing linear hypotheses" - but with a slight twist. Our large model is still the full regression model (from the fitted object), but the smaller model is replaced by the _the change from one model to the next_.

Let SSE be the sums-of-squares-error (residual sums of squares) from the full (large, called A) model - this will be our denominator (as always).
For our rent example the denominator will be SSE/(n-p)=64819547/3075 (see above).

The logic is that the full model provides an estimate of $\sigma^2$: the others may not.

---

For the numerator we are not comparing one small model with the full (large) one, we are instead looking at the change in SSE between two (smaller) models (called model B1 and B2). So, now we have in the numerator the difference in SSE between models B1 and B2, scaled with the difference in number of parameters estimated in model B1 and B2 ="number in B2 minus in B1" (which is the same as the difference in degrees of freedom for the two models).

---

This means that the test statistics we use are:

$$ F_0=\frac{\frac{\text{SSE}_{B1}-\text{SSE}_{B2}}{\text{df}_{B1}-\text{df}_{B2}}}{\frac{\text{SSE}_A}{\text{df}_A}}$$
Remark: notice that the denominator is just the $\hat{\sigma^2}$ from the larger model A.

This makes our $F$-test statistic: $f_0=\frac{40299098/1}{64819547/3075}=1911.765$ (remember that we swap from capital to small letters when we insert numerical values).

To produce a $p$-value to the test that
$$H_0: \text{"Model B1 and B2 are equally good" vs }H_1:\text{"Model B2 is better than B1}$$
and then the $F\sim {\text{df}_{B1}-\text{df}_{B2},\text{df}_A}$. 

---

In our example we compare to an F-distribution with 1 and 3075 degrees of freedom. The $p$-value is the "probability of observing a test statistic at least as extreme as we have" so we calculate the $p$-value as $P(F>f_0)$. This gives a $p$-value that is practically 0.

If you then want to use the asymptotic version (relating to a chi-square instead of the F), then multiply your F-statistic with $\text{df}_{B1}-\text{df}_{B2}$ and relate to a $\chi^2$ distribution with $\text{df}_{B1}-\text{df}_{B2}$ degrees of freedom, where $\text{df}_{B1}-\text{df}_{B2}$ is the difference in number of parameters in models B1 and B2. In our example $\text{df}_{B1}-\text{df}_{B2}=1$. 

---

For the anova table we do this sequentially for all models from starting with only intercept to the full model A. This means you need to calculate SSE and df for models of all sizes to calculate lots of these $F_0$s. Assume that we have 4 covariates that are added to the model, and call the 5 possible models (given the order of adding the covariates) 

* model 1: model with only intercept
* model 2: model with intercept and covariate 1
* model 3: model with intercept and covariate 1 and covariate 2
* model 4: model with intercept and covariate 1 and covariate 2 and covariate 3
* model 5: model with intercept and covariate 1 and covariate 2 and covariate 3 and covariate 4

---

Fit a linear model (lm) for each model 1-5, and store SSE and degrees of freedom=df (number of observations minus number of covariates estimated) for each of the models. Call these SSE$_1$ to SSE$_5$ and df$_1$ to $df$_5$.

The anova output has columns: Df   Sum Sq  Mean Sq  F value    Pr(>F) 
and one row for each covariate added to the model.

For example 

model 2 vs model 1: Df=df$_1$-df$_2$, Sum Sq=SSE$_1$-SSE$_2$, Mean Sq=Sum Sq/Df, F value=(Mean Sq)/(SSE$_5$/df$_5$)=$f_0$, Pr(>F)=pvalue=$P(F>f_0)$.

model 3 vs model 2: Df=df$_2$-df$_3$, Sum Sq=SSE$_2$-SSE$_3$, Mean Sq=Sum Sq/Df, F value=(Mean Sq)/(SSE$_5$/df$_5$)=$f_0$, Pr(>F)=pvalue=$P(F>f_0)$.


In R the p-value is calculated as `1-pf(f0,Df)` or as `1-pchisq(Df*f0,Df)` if the asymptotic chisquare distribution is used.

This is what is presented - a sequential record of the effect of adding a new covariate.

---

**Q*: What if you change the order of the covariates into the model? 

<!---
Yes, then everything changes. That is the drawback of Type I (sequential) thinking.
--->

---

A competing way of thinking is called _type 3 ANOVA_ and instead of looking sequentially at adding terms, we (like in `summary`) calculated the contribution to a covariate (or factor) given that all other covariates are present in the regression model. Type 3 ANOVA is available from library `car` as function `Anova` (possible to give type of anova as input).

**Check** : Take a look at the print-out from  `summary` and  `anova` and observe that for our rent data the $p$-values for each covariate are different due to the different nature of the $H_0$s tested (sequential vs. "all other present").

If we had orthogonal columns for our different covariates the type 1 and type 3 ANOVA tables would have been equal.


---

<!---
#### Optional (beyond the scope of this course)

There is also something called a type 2 ANOVA table, but that is mainly important if we have interactions in our model, so we do not consider that here. If you want to read more this blogplot
<http://goanna.cs.rmit.edu.au/~fscholer/anova.php>
is a good read. And, in combination with different variants of dummy and effct coding - read this:
<http://rstudio-pubs-static.s3.amazonaws.com/65059_586f394d8eb84f84b1baaf56ffb6b47f.html>

--->

## Quality measures

To assess the quality of the regression we can report the $R^2$ coefficient of determination. However, since adding covariates to the linear regression can not make the SSE larger, this means that adding covariates can not make the $R^2$ smaller. This means that SSE and $R^2$ are only useful measures for comparing models with the same number of regression parameters estimated.

If we consider two models with the same model complexity then SSE can be used to choose between (or compare) these models.

But, if we want to compare models with different model complexity we need to look at other measures of quality for the regression.

---

### $R^2$ adjusted (corrected) 
$$R^2_{\text{adj}}=1-\frac{\frac{SSE}{n-p}}{\frac{SST}{n-1}}=1-\frac{n-1}{n-p}(1-R^2)$$
Choose the model with the _largest_ $R^2_{\text{adj}}$.

----

### AIC Akaike information criterion 

AIC is one of the most widely used criteria, and is designed for likelihood-based inference.
Let $l(\hat{\beta}_M,\tilde{\sigma}^2)$ be the maximum of the log-likelihood of the data inserted the maximum likelihood estimates for the regression and nuisance parameter. Further, let $\lvert M \rvert$ be the number of estimated regression parameters in our model.
$$\text{AIC} =-2 \cdot l(\hat{\beta}_M,\tilde{\sigma}^2)+2(\lvert M\rvert +1)$$

For a normal regression model:
$$\text{AIC} =n\ln(\tilde{\sigma}^2)+2(\lvert M\rvert +1)+C$$
where C is a function of $n$ (will be the same for two models for the same data set).
Remark that $\tilde{\sigma}^2=SSE/n$ - our ML estimator (not our unbiased REML), so that the first term in the AIC is just a function of the SSE. For MLR the AIC and the Mallows Cp gives the same result when comparing models.

Choose the model with the minimum AIC. 

---

### BIC  Bayesian information criterion.

The BIC is also based on the likelihood (see notation above).
$$\text{BIC} =-2 \cdot l(\hat{\beta}_M,\tilde{\sigma}^2)+\ln(n)\cdot (\lvert M\rvert +1)$$

For a normal regression model:
$$ \text{BIC}= n\ln(\tilde{\sigma}^2)+\ln(n)(\lvert M\rvert +1)$$
Choose the model with the minimum BIC. 

AIC and BIC are motivated in very different ways, but the final result for the normal regression model is very similar.
BIC has a larger penalty than AIC ($\log(n)$ vs. $2$), and will often give a smaller model (=more parsimonious models) than AIC. 
In general we would not like a model that is too complex.

---

## Model selection strategies

* All subset selection: use smart "leaps and bounds" algorithm, works fine for number of covariates in the order of 40.

* Forward selection: choose starting model (only intercept), then add one new variable at each step - selected to make the best improvement in the model selection criteria. End when no improvement is made.

* Backward elimination: choose starting model (full model), then remove one new variable at each step - selected to make the best improvement in the model selection criteria. End when no improvement is made.

* Stepwise selection: combine forward and backward.

---

# R packages

```{r, eval=FALSE}
install.packages(c("gamlss.data","tidyverse", "GGally", 
                   "Matrix", "nortest"))
```

# References and further reading

* Slightly different presentation (more focus on multivariate normal theory): [Slides and written material from TMA4267 Linear Statistical Models in 2017, Part 2:  Regression (by Mette Langaas).](https://www.math.ntnu.no/emner/TMA4267/2017v/TMA4267V2017Part2.pdf)
* And, same source, but now [Slides and written material from TMA4267 Linear Statistical Models in 2017, Part 3: Hypothesis testing and ANOVA] (http://www.math.ntnu.no/emner/TMA4267/2017v/TMA4267V2017Part3.pdf)


